---
title: "Data_Acquisition"
output: html_document
date: "2026-09-21"
editor_options: 
  markdown: 
    wrap: 72
---

# Library-Readin

Notes:

(09/24/26): I forgot that the clinical data is also needed

```{r}
library(Seurat)
library(tidyverse)
library(ggplot2)
library(hdf5r)
library(harmony)
#library(BPCells)
library(SeuratWrappers)
library(GEOquery)
```

[hackathon project](https://github.com/mitreac/BIDS2026_projects)

Requirements for Data Acqusition:

Prostate Cancer or Adj Normal (Ideally split even)

Paired with Clinical Data

During Acquisition, Normal and Tumor is required for finding low and
high risk profiles Dataset will have to consist of 6PRAD-SC (preferably
10x Chromium) 3Normal-Adj, and 3PRAD datasets. Integrate? Yes
integration via CCA would be suffient to compare transcriptional markers
and hopefully have a gradient of cellullar change from Low to High risk
factors. However within the each meta dat source columns will be very
important and I may have to import this on the windows PC.

Status Update (09/26/2026): There is a lack of clinical information on
the Single-Cell RNAseq side (GEOACCESSION, TISCH,, BROAD). However there
is an abundance on the bulkRNA-SEQ side. Which means how can I utilize
the abundance of clinical data with PRAD bulkRNA-Seq data in tandem with
the resolution of single cell PRAD data?

Goal: Finding a trend between immune cell landscape and clinical status
using scRNASeq and bulkRNA-Seq data

Papers that could help:

[https://pmc.ncbi.nlm.nih.gov/articles/PMC12161480](https://pmc.ncbi.nlm.nih.gov/articles/PMC12161480/?utm_source=chatgpt.com){.uri}

<https://pubmed.ncbi.nlm.nih.gov/34936871/>

<https://pmc.ncbi.nlm.nih.gov/articles/PMC11369112/?utm_source=chatgpt.com>

# Datasets Available

### PRAD Cancer Sample

- [GSE141445 (Chen et. al., 2021)
  (PRAD)](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE141445)

- [GSE176031 (Sang, H 2021)
  (PRAD)](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE176031)

- [GSE150692(Crowly, L.,
  2020)](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE150692)(PRAD)

### Paired/ Adj.N Datasets

- [EGAD01008340 (Tuang et. al, 2021)
  (Paird-PRAD-Adj.N)](https://ega-archive.org/datasets/EGAD00001008340)

- [GSE181294(Chen et. al., 2023)
  (Paird-PRAD-Adj.N)](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE181294)

- [GSE1201716(Henry et. al., 2018)
  (Adj.N)](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE120716)

+--------------+----------------------------------------------------------------------+-------------+--------------+----------+-----------+---------------+--------------+
|              | Author(s)                                                            | Organism(s) | Condition(s) | Cell Cnt | Lib Prep  | Availability  | Clinical     |
|              |                                                                      |             |              |          | Protocol  | Format        | Data         |
+==============+======================================================================+=============+==============+==========+===========+===============+==============+
| GSE141445    | [(Chen et. al.,                                                      | Human       | PRAD         | 36,424   | 10x       | series_matrix | available,   |
|              | 2021)](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE141445) |             |              |          | Chromium  |               | but sparse   |
|              |                                                                      |             |              |          | V2        |               |              |
+--------------+----------------------------------------------------------------------+-------------+--------------+----------+-----------+---------------+--------------+
| GSE176031    | [(Song, H                                                            | Human       | Paired       | 21,743   | Seq-Well  | series_matrix | available,   |
|              | 2021)](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE176031) |             | (AdjNorm +   |          |           |               | but sparse   |
|              |                                                                      |             | PRAD)        |          |           |               | and          |
|              |                                                                      |             |              |          |           |               | inconsistent |
+--------------+----------------------------------------------------------------------+-------------+--------------+----------+-----------+---------------+--------------+
| GSE150692    | [(Crowley,                                                           | Human / Mus | PRAD         |          | Chromium  | series_matrix | sparse but   |
|              | 2020)](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE150692) | Musculus    |              |          | Single    |               | available    |
|              |                                                                      |             |              |          | Cell 5’   |               |              |
|              |                                                                      | "orthotopic |              |          | Kit       |               | only 1 out   |
|              |                                                                      | 6694c2      |              |          |           |               | of three has |
|              |                                                                      | pancreatic  |              |          | DO NOT    |               | useful       |
|              |                                                                      | tumors"     |              |          | USE       |               | clinical     |
|              |                                                                      |             |              |          |           |               | data         |
+--------------+----------------------------------------------------------------------+-------------+--------------+----------+-----------+---------------+--------------+
| EGAD01008340 | [(Tuang et. al,                                                      | Human       | Paired (PRAD |          |           | .fast_files   | not          |
|              | 2021)](https://ega-archive.org/datasets/EGAD00001008340)             |             | / N. Adj)    |          |           |               | available    |
+--------------+----------------------------------------------------------------------+-------------+--------------+----------+-----------+---------------+--------------+
| GSE181294    | [(](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi)[Chen et. al.,    | Human       | Paired (PRAD |          | slide-seq | series matrix | available:   |
|              | 2023)](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE181294) |             | / N. Adj)    |          |           |               |              |
|              |                                                                      |             |              |          |           |               | condition,   |
|              |                                                                      |             |              |          |           |               | age,         |
|              |                                                                      |             |              |          |           |               | gleason,     |
|              |                                                                      |             |              |          |           |               | grade,       |
|              |                                                                      |             |              |          |           |               | tissue       |
+--------------+----------------------------------------------------------------------+-------------+--------------+----------+-----------+---------------+--------------+
| GSE120716    | [(Henry et. al.,                                                     | Human       | N. Adj       |          |           | series matrix |              |
|              | 2018)](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE120716) |             |              |          |           | file          |              |
+--------------+----------------------------------------------------------------------+-------------+--------------+----------+-----------+---------------+--------------+

# GSE141445/Chen, O Download

The 13 prostate tumor samples used in this study were collected from
patients who had undergone radical prostatectomy at Shanghai Changhai
Hospital. Single cell RNA sequencing was performed with 10XGenomics.

MetaData Columns: (orig.ident, sample_title, disease state, patientID
(if possible))

Seq-Well S\^3 protocol

```{r}
library(GEOquery)
#enter in GSE141445
#read in the whole count matrices and meta data

gse_141445 <- getGEO("GSE141445", GSEMatrix = TRUE)

#divide the ExpressionSet into the expression matrix and meta data

gse_141445 <- gse_141445[[1]]

#expression matrix

expr_141445 <- exprs(gse_141445)

#meta-data
meta_141445 <- pData(gse_141445)

#showing first five rows of the meta-data and the expression samples
head(expr_141445, n = 5)
head(meta_141445, n = 5)
#download the data into the data directory
getGEOSuppFiles("GSE141445")
```

```{r}
meta_141445
```

```{r, inlcude=FALSE, warning=FALSE}
library(glue)
dir <- "GSE141445"
filename <- "GSM4203181_data.raw.matrix.txt.gz"

if (!file.exists(dir)) {
  dir.create(dir, showWarnings = FALSE)
}
# Huge Reason why I am using this instead of regular GEOquery
# the GSEXXXXX_RAW.tar that came form GEO,query only comes with
#
system(glue(
  "if [ ! -f ./{dir}/{filename} ]; then ",
  "wget -P ./{dir}/ ",
  "https://ftp.ncbi.nlm.nih.gov/geo/samples/GSM4203nnn/GSM4203181/suppl/",
  "{filename}; ",
  "fi"
))
```

### Create Seurat Object and Edit MetaData

```{r}
library(glue)
#read in via path name from downloaded file
path <- glue("./{dir}/{filename}")
counts <- read.delim(gzfile(path))
cnt_mat <- as.matrix(counts)
rm(counts)
gse141445.obj <- CreateSeuratObject(cnt_mat)
rm(cnt_mat)
gse141445.obj@meta.data$orig.ident <- rep("Chen2021", nrow(gse141445.obj@meta.data))

gse141445.obj@meta.data$gse_acc <- rep("GSE141445", nrow(gse141445.obj@meta.data))

gse141445.obj@meta.data$condition <- rep("tumor", nrow(gse141445.obj@meta.data))
```

### Saving GSE141445 as an RDS Object

```{r}
dir <- "sc_rds_files"
if (!file.exists(dir)) {
  dir.create(dir, showWarnings = FALSE)
}
saveRDS(gse141445.obj, glue("./{dir}/gse141445.rds"))
#removing non essential data from the

rm(gse_141445,meta_141445,expr_141445)
```

```{r}
gse141445.obj <- readRDS("./sc_rds_files/gse141445.rds")
```

# GSE176031/Song, H Download

Single-cell RNA sequencing based on Seq-Well S\^3 protocol of **6
prostate biopsies from 3 different patients, 4 radical prostatectomies
with tumor-only samples from 4 patients, and 4 radical prostatectomies
with matched normal samples from 4 patients**. Organoids derived from
radical prostatectomy paired tumor and normal tissue samples.

MetaData columns:

**MetaData Columns: (orig.ident, sample_title, disease state, patientID
(if possible))**

```{r}
library(GEOquery)
#enter in GSE141445
#read in the whole count matrices and meta data

gse_176031 <- getGEO("GSE176031", GSEMatrix = TRUE)
#divide the ExpressionSet into the expression matrix and meta data

gse_176031 <- gse_176031[[1]]

#expression matrix

expr_176031 <- exprs(gse_176031)

#meta-data
meta_176031 <- pData(gse_176031)

#showing first five rows of the meta-data and the expression samples
head(expr_176031, n = 5)
head(meta_176031, n = 5)
#download the data into the data directory
getGEOSuppFiles("GSE176031")
```

```{r}
meta_176031
```

```{r}
meta_176031 <- meta_176031 %>% 
  select(c(title,geo_accession, source_name_ch1,"patient:ch1", "tissue:ch1" ))
```

```{r}
state_markers <- c(unique(meta_176031$source_name_ch1))
marker_replacements <- c("biopsy","tumor","norm_adj","tumor_adj","organoid")
state_markers
```

```{r}
#Intent is to replace the column's entries in the first character vector
# with the entries in the character vector
state_markers <- c(unique(meta_176031$source_name_ch1))
marker_replacements <- c("biopsy","tumor","norm_adj","tumor_adj","organoid")

replacement <- setNames(marker_replacements, state_markers)

meta_176031 <- meta_176031 %>%
  mutate(
    source_name_ch1 = replacement[source_name_ch1]
  ) %>%
  select(c(geo_accession, source_name_ch1)) %>%
  mutate(geo_source = paste0(geo_accession,"_", source_name_ch1))
```

```{r}
meta_176031
```

```{r, include=FALSE}
library(glue)

dir <- "GSE176031"
ftp <- "https://ftp.ncbi.nlm.nih.gov/geo/series/GSE176nnn/GSE176031/matrix/"
filename <- "GSE176031_RAW.tar"

if (!file.exists(dir)) {
  dir.create(dir, showWarnings = FALSE)
}
# Huge Reason why I am using this instead of regular GEOquery
# the GSEXXXXX_RAW.tar that came form GEO,query only comes with
#
system(glue(
  "if [ ! -f ./{dir}/{filename} ]; then ",
  "wget -P ./{dir}/ ",
  "{ftp}",
  "{filename}; ",
  "fi"
))

system(glue("tar -xzf ./{dir}/{filename} -C ./{dir}/"))
```

```{r}

files <- list.files(path = "./GSE176031",
                    pattern = "\\_dge.txt.gz",
                    recursive = TRUE,
                    full.names = TRUE)
basename(files)[1:3]

cnt_mats_176031 <- list()
for(i in seq_along(files)){
  cnt_mats_176031[[i]] <- read.delim(gzfile(files[i]))
  
  cnt_mats_176031[[i]] <- column_to_rownames(
    cnt_mats_176031[[i]],
    "GENE"
  )
}
names(cnt_mats_176031) <- sub("^(GSM[0-9]+_).*", "\\1", basename(files))
names(cnt_mats_176031) <- sub("_", "", names(cnt_mats_176031))

names(cnt_mats_176031) <- meta_176031$geo_source
```

### Barcode Editing

```{r}
common_features_176031 <- Reduce(intersect, lapply(cnt_mats_176031, rownames))


cnt_mats_common_176031 <- lapply(
  cnt_mats_176031,
  \(x) x[common_features_176031, ,drop = FALSE]
)
# for some reason this is appending the list names onto the colnames
 # Which give global uniqueness to the barcodes
#However take that away, and we have non-unique barcodes between sequencing libraries
#I need to figure out why the names(cnt_mats) gets pre-pended to the DNA barcodes
count_matrix_176031 <- do.call(cbind, cnt_mats_common_176031)
#rm(cnt_mats)
#rm(cnt_mats_common)
length(unique(rownames(count_matrix_176031))) == nrow(count_matrix_176031)
```

Checking globalm uniquenes not inter-library uniqueness

```{r}
#rm(cnt_mats)
#rm(cnt_mats_common)
count_matrix_176031 <- as.matrix(count_matrix_176031)
length(unique(rownames(count_matrix_176031))) == nrow(count_matrix_176031)

#There is global unique-ness between the barcodes, but not uniqueness between libraries
anyDuplicated(colnames(count_matrix_176031))
any(is.na(colnames(count_matrix_176031)))
any(colnames(count_matrix_176031) == "")
```

### Editing Seurat Meta-Data

```{r}
gse176031.obj <- CreateSeuratObject(count_matrix_176031)
gse176031.obj@meta.data$orig.ident <- "Song2021"
gse176031.obj@meta.data$geo_acc <- "GSE176031"

cell_names <-  rownames(gse176031.obj@meta.data)

sample_title <- sub("\\..*", "", cell_names)

gse176031.obj@meta.data$sample_title  <- sample_title

# this will address the inter library non-unique scenario
non_unique_barcodes <- sub("^[^.]*\\.", "", cell_names)

n <- ave( non_unique_barcodes, non_unique_barcodes, FUN = seq_along)

result <- paste0(non_unique_barcodes, ".", n)

rownames(gse176031.obj@meta.data) <- result

gse176031.obj@meta.data <- gse176031.obj@meta.data %>%
  mutate(condition = str_extract(sample_title, "(?<=_).*")) %>%
  mutate(sample = str_extract(sample_title, "^[^_]+")) %>%
  select(-sample_title)

saveRDS(gse176031.obj, "./sc_rds_files/gse176031.rds")

```

```{r}
gse176031.obj@meta.data
```

```{r}
saveRDS(gse176031.obj, "./sc_rds_files/gse176031.rds")
```

```{r}
rm(cnt_mats_176031, cnt_mats_common_176031, count_matrix_176031, meta_176031, gse176031.obj)
```

Problem: The authors stated that they had \~21,000 cells after QC.
However after concatenating we have an insane 53,00 cells. It's very
likely these are not QCed but it's a concerning amount to be that far
from the Paper's numbers

We'll have to see after QC if this reduces at all highly likely it will
reduce these cells by alot

# [GSE150692 (Crowley, 2020) (PRAD](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE150692)) Download

2 normal benign prostate tissue specimens obtained from patient
resections.

**10x Genomics Chromium platform**

In the Supplementary Data Crowley, et.al state that they used 6 human
prostate samples, but only 3 are availble for download in the GEOquery
website. BRUH

**Human Samples Available on GEOQuery:**

GSM4556600(MF002) clinical data available, benign tumor

GSM4556601(MM033), benign tumor

GSM4556602(MM037) , benign tumor

**Mouse Samples (GeoQuery MetaData only gives the GSMXXXs for mouse
Samples):**

GSM4556594 GSM4556595 GSM4556596 GSM4556597 GSM4556598 GSM4556599

DO NOT INCLUDE MOUSE SAMPLES LOL

```{r}
library(GEOquery)
#enter in GSE141445
#read in the whole count matrices and meta data

gse_150692 <- getGEO("GSE150692", GSEMatrix = TRUE)
#divide the ExpressionSet into the expression matrix and meta data

gse_150692 <- gse_150692[[1]]

#expression matrix

expr_150692 <- exprs(gse_150692)

#meta-data
meta_150692 <- pData(gse_150692)

#showing first five rows of the meta-data and the expression samples
head(expr_150692, n = 5)
head(meta_150692, n = 5)
#download the data into the data directory
getGEOSuppFiles("GSE150692")
```

```{r}
meta_150692
```

### GEOquery Pull Excludes Human Assays

```{r}
library(glue)
dir <- "GSE150692"
filename <- "GSE150692_RAW.tar"

if (!file.exists(dir)) {
  dir.create(dir, showWarnings = FALSE)
}
# Huge Reason why I am using this instead of regular GEOquery
# the GSEXXXXX_RAW.tar that came form GEO,query only comes with
#
system(glue(
  "if [ ! -f ./{dir}/{filename} ]; then ",
  "wget -P ./{dir}/ ",
  "https://ftp.ncbi.nlm.nih.gov/geo/series/GSE150nnn/GSE150692/suppl/",
  "{filename}; ",
  "fi"
))

system(glue("tar -xzf ./{dir}/{filename} -C ./{dir}/"))

#GSMXX60 Are all human samples, but

system(glue(
  "mv ./{dir}/GSM4556602_MM037_clean_raw_counts_matrix.txt.gz ",
  "./{dir}/GSM4556602_MM037_human_clean_raw_counts_matrix.txt.gz"
))

#sweep for only human files in the downloaded data
files <- list.files(path = "./GSE150692",
                    pattern = "_human_.*\\.txt\\.gz$",
                    recursive = TRUE,
                    full.names = TRUE)
                    
basename(files)

cnt_mats_150692 <- list()
for(i in seq_along(files)){
  cnt_mats_150692[[i]] <- read.delim(gzfile(files[i]))
}



names(cnt_mats_150692) <- sub("^(GSM[0-9]+_).*", "\\1", basename(files))
names(cnt_mats_150692) <- sub("_", "", names(cnt_mats_150692))
```

```{r}
common_features <- Reduce(intersect, lapply(cnt_mats_150692, function(x) x$X))

common_features[1:20]
```

### Fixing Repeated Gene Names in the data (Warning, Very Long Wait)

```{r}

cnt_mats_common_150692 <- lapply(cnt_mats_150692, function(x) {
  x %>%
    group_by(X) %>%
    summarise(across(where(is.numeric), sum)) %>%
    filter(X %in% common_features)
})
# for some reason this is appending the list names onto the colnames
# Which give global uniqueness to the barcodes
#However take that away, and we have non-unique barcodes between sequencing libraries
#I need to figure out why the names(cnt_mats) gets pre-pended to the DNA barcodes
count_matrix_150692 <- do.call(cbind, cnt_mats_common_150692)
```

```{r}
count_matrix_150692 <- as.tibble(count_matrix_150692) %>% column_to_rownames("GSM4556600.X")
```

```{r}
rm(cnt_mats_150692)
rm(cnt_mats_common_150692)
count_matrix <- as.matrix(count_matrix_150692)
length(unique(rownames(count_matrix_150692))) == nrow(count_matrix_150692)

#There is global unique-ness between the barcodes, but not uniqueness between libraries
anyDuplicated(colnames(count_matrix_150692))
any(is.na(colnames(count_matrix_150692)))
any(colnames(count_matrix_150692) == "")
```

```{r}
gse150692.obj <- CreateSeuratObject(count_matrix_150692)
```

Editing Meta Data

Human samples only from GSE150692

```{r}
gse150692.obj

gse150692.obj@meta.data$orig.ident <- "Crowley2020"
gse150692.obj@meta.data$geo_acc <- "GSE150692"

cell_names <-  rownames(gse150692.obj@meta.data)

sample_title <- sub("\\..*", "", cell_names)

gse150692.obj@meta.data$sample_title  <- sample_title

# this will address the inter library non-unique scenario
# this will address the inter library non-unique scenario
non_unique_barcodes <- sub("^[^.]*\\.", "", cell_names)

n <- ave(non_unique_barcodes, non_unique_barcodes, FUN = seq_along)

result <- paste0(non_unique_barcodes, ".", n)

rownames(gse150692.obj@meta.data) <- result

```

```{r}
gse150692.obj@meta.data
#saveRDS(gse150692.obj, "./sc_rds_files/gse150692.rds")
```

# GSE181294 ([Chen et. al., 2023](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE181294))

Sequenced with Illumina NextSeq 500, Prepped with 10x Chromium Protocol
(Single Cell 3' Library and Gel Bead Kit v2)

20 human prostate tumors + matched adj Normal tissues

5 Additional Healthy Prostates

```{r}
library(GEOquery)
#enter in GSE141445
#read in the whole count matrices and meta data

gse_181294 <- getGEO("GSE181294", GSEMatrix = TRUE)
#divide the ExpressionSet into the expression matrix and meta data

gse_181294 <- gse_181294[[1]]

#expression matrix

expr_181294 <- exprs(gse_181294)

#meta-data
meta_181294 <- pData(gse_181294)

#showing first five rows of the meta-data and the expression samples
head(expr_181294, n = 5)
head(meta_181294, n = 5)
#download the data into the data directory
getGEOSuppFiles("GSE181294")
```

```{r}
meta_181294
```

```{r}
colnames(meta_181294)
meta_181294 <- meta_181294 %>% select(c(title,
                                        geo_accession,
                                        "age:ch1",
                                        "condition:ch1",
                                        "gleason:ch1",
                                        "grade:ch1",
                                        "tissue:ch1"))
meta_181294
```

```{r}
library(glue)
dir <- "GSE181294"
filename <- "GSE181294_RAW.tar"

if (!file.exists(dir)) {
  dir.create(dir, showWarnings = FALSE)
}
# Huge Reason why I am using this instead of regular GEOquery
# the GSEXXXXX_RAW.tar that came form GEO,query only comes with
#
system(glue(
  "if [ ! -f ./{dir}/{filename} ]; then ",
  "wget -P ./{dir}/ ",
  "https://ftp.ncbi.nlm.nih.gov/geo/series/GSE181nnn/GSE181294/suppl/",
  "{filename}; ",
  "fi"
))

system(glue("tar -xzf ./{dir}/{filename} -C ./{dir}/"))
```

```{r}
files <- list.files(path = "./GSE181294",
                    pattern = "_human_.*\\.txt\\.gz$",
                    recursive = TRUE,
                    full.names = TRUE)
basename(files)[1:2]

cnt_mats_150692 <- list()
for(i in seq_along(files)){
  cnt_mats_150692[[i]] <- read.delim(gzfile(files[i]))
}

names(cnt_mats_150692) <- sub("^(GSM[0-9]+_).*", "\\1", basename(files))
names(cnt_mats_150692) <- sub("_", "", names(cnt_mats_150692))
```

```{r}

```

# GSE120716 [(Henry et. al., 2018)](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE120716)

GSE1207816 is a SuperSeries of two different resolutions of RNA-Seq data

[GSE117271](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE117271):
Bulk RNAseq data of cell types isolated by FACS from normal human
prostates

[GSE117403](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE117403):
scRNA-Seq of cell types isolated by FACS from normal human prostates

```{r}
library(GEOquery)
#enter in GSE141445
#read in the whole count matrices and meta data

gse_117403 <- getGEO("GSE117403", GSEMatrix = TRUE)
#divide the ExpressionSet into the expression matrix and meta data

gse_117403 <- gse_117403[[1]]

#expression matrix

expr_117403 <- exprs(gse_117403)

#meta-data
meta_117403 <- pData(gse_117403)

#showing first five rows of the meta-data and the expression samples
head(expr_117403, n = 5)
head(meta_117403, n = 5)
#download the data into the data directory
getGEOSuppFiles("GSE117403")
```

```{r}

```

```{r}
library(glue)
dir <- "GSE117271"
filename <- "GSE117271_series_matrix.txt.gz"

if (!file.exists(dir)) {
  dir.create(dir, showWarnings = FALSE)
}
# Huge Reason why I am using this instead of regular GEOquery
# the GSEXXXXX_RAW.tar that came form GEO,query only comes with
#
system(glue(
  "if [ ! -f ./{dir}/{filename} ]; then ",
  "wget -P ./{dir}/ ",
  "https://ftp.ncbi.nlm.nih.gov/geo/series/GSE117nnn/GSE117271/matrix/",
  "{filename}; ",
  "fi"
))

path <- glue("./{dir}/{filename}")
counts <- read.delim(gzfile(path))
cnt_mat <- as.matrix(counts)
```
