---
title: "Data_Acquisition"
output: html_document
date: "2026-09-21"
editor_options: 
  markdown: 
    wrap: 72
---

# Library-Readin

Notes:

```{r}
library(Seurat)
library(tidyverse)
library(ggplot2)
library(hdf5r)
library(harmony)
#library(BPCells)
library(SeuratWrappers)
library(GEOquery)
```

[hackathon project](https://github.com/mitreac/BIDS2026_projects)

Requirements for Data Acqusition:

Prostate Cancer or Adj Normal (Ideally split even)

Paired with Clinical Data

During Acquisition, Normal and Tumor is required for finding low and
high risk profiles Dataset will have to consist of 6PRAD-SC (preferably
10x Chromium) 3Normal-Adj, and 3PRAD datasets. Integrate? Yes
integration via CCA would be suffient to compare transcriptional markers
and hopefully have a gradient of cellullar change from Low to High risk
factors. However within the each meta dat source columns will be very
important and I may have to import this on the windows PC.

Status Update (09/26/2026): There is a lack of clinical information on
the Single-Cell RNAseq side (GEOACCESSION, TISCH,, BROAD). However there
is an abundance on the bulkRNA-SEQ side. Which means how can I utilize
the abundance of clinical data with PRAD bulkRNA-Seq data in tandem with
the resolution of single cell PRAD data?

Goal: Finding a trend between immune cell landscape and clinical status
using scRNASeq and bulkRNA-Seq data

Papers that could help:

[https://pmc.ncbi.nlm.nih.gov/articles/PMC12161480](https://pmc.ncbi.nlm.nih.gov/articles/PMC12161480/?utm_source=chatgpt.com){.uri}

<https://pubmed.ncbi.nlm.nih.gov/34936871/>

<https://pmc.ncbi.nlm.nih.gov/articles/PMC11369112/?utm_source=chatgpt.com>

# Proposed Datasets Available

### PRAD Cancer Sample

- [GSE141445 (Chen et. al., 2021)
  (PRAD)](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE141445)

- [GSE176031 (Sang, H 2021)
  (PRAD)](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE176031)

- [GSE150692(Crowly, L.,
  2020)](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE150692)(PRAD)

### Paired/ Adj.N Datasets

- [EGAD01008340 (Tuang et. al, 2021)
  (Paird-PRAD-Adj.N)](https://ega-archive.org/datasets/EGAD00001008340)

- [GSE181294(Chen et. al., 2023)
  (Paird-PRAD-Adj.N)](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE181294)

- [GSE1201716(Henry et. al., 2018)
  (Adj.N)](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE120716)

+--------------+----------------------------------------------------------------------+-------------+--------------+----------+-----------+---------------+--------------+
|              | Author(s)                                                            | Organism(s) | Condition(s) | Cell Cnt | Lib Prep  | Availability  | Clinical     |
|              |                                                                      |             |              |          | Protocol  | Format        | Data         |
+==============+======================================================================+=============+==============+==========+===========+===============+==============+
| GSE141445    | [(Chen et. al.,                                                      | Human       | PRAD         | 36,424   | 10x       | series_matrix | available,   |
|              | 2021)](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE141445) |             |              |          | Chromium  |               | but sparse   |
|              |                                                                      |             |              |          | V2        |               |              |
+--------------+----------------------------------------------------------------------+-------------+--------------+----------+-----------+---------------+--------------+
| GSE176031    | [(Song, H                                                            | Human       | Paired       | 21,743   | Seq-Well  | series_matrix | available,   |
|              | 2021)](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE176031) |             | (AdjNorm +   |          |           |               | but sparse   |
|              |                                                                      |             | PRAD)        |          |           |               | and          |
|              |                                                                      |             |              |          |           |               | inconsistent |
+--------------+----------------------------------------------------------------------+-------------+--------------+----------+-----------+---------------+--------------+
| GSE150692    | [(Crowley,                                                           | Human / Mus | PRAD         |          | Chromium  | series_matrix | sparse but   |
|              | 2020)](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE150692) | Musculus    |              |          | Single    |               | available    |
|              |                                                                      |             |              |          | Cell 5’   |               |              |
|              |                                                                      | "orthotopic |              |          | Kit       |               | only         |
|              |                                                                      | 6694c2      |              |          |           |               |              |
|              |                                                                      | pancreatic  |              |          | DO NOT    |               | 1 out of     |
|              |                                                                      | tumors"     |              |          | USE       |               | three has    |
|              |                                                                      |             |              |          |           |               | useful       |
|              |                                                                      |             |              |          |           |               | clinical     |
|              |                                                                      |             |              |          |           |               | data.        |
|              |                                                                      |             |              |          |           |               |              |
|              |                                                                      |             |              |          |           |               | Not worth    |
|              |                                                                      |             |              |          |           |               | trying to    |
|              |                                                                      |             |              |          |           |               | use the      |
|              |                                                                      |             |              |          |           |               | clinical     |
|              |                                                                      |             |              |          |           |               | data         |
+--------------+----------------------------------------------------------------------+-------------+--------------+----------+-----------+---------------+--------------+
| EGAD01008340 | [(Tuang et. al,                                                      | Human       | Paired (PRAD |          |           | .fast_files   | not          |
|              | 2021)](https://ega-archive.org/datasets/EGAD00001008340)             |             | / N. Adj)    |          |           |               | available    |
+--------------+----------------------------------------------------------------------+-------------+--------------+----------+-----------+---------------+--------------+
| GSE181294    | [(](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi)[Chen et. al.,    | Human       | Paired (PRAD |          | slide-seq | series matrix | available:   |
|              | 2023)](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE181294) |             | / N. Adj)    |          |           |               |              |
|              |                                                                      |             |              |          |           | Not possible  | condition,   |
|              |                                                                      |             |              |          |           | too big even  | age,         |
|              |                                                                      |             |              |          |           | for the big   | gleason,     |
|              |                                                                      |             |              |          |           | computer      | grade,       |
|              |                                                                      |             |              |          |           |               | tissue       |
+--------------+----------------------------------------------------------------------+-------------+--------------+----------+-----------+---------------+--------------+
| GSE120716    | [(Henry et. al.,                                                     | Human       | N. Adj       |          |           | series matrix |              |
|              | 2018)](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE120716) |             |              |          |           | file          |              |
+--------------+----------------------------------------------------------------------+-------------+--------------+----------+-----------+---------------+--------------+

# GSE141445/Chen, O Download

The 13 prostate tumor samples used in this study were collected from
patients who had undergone radical prostatectomy at Shanghai Changhai
Hospital. Single cell RNA sequencing was performed with 10XGenomics.

MetaData Columns: (orig.ident, sample_title, disease state, patientID
(if possible))

Seq-Well S\^3 protocol

```{r}
library(GEOquery)
#enter in GSE141445
#read in the whole count matrices and meta data

gse_141445 <- getGEO("GSE141445", GSEMatrix = TRUE)

#divide the ExpressionSet into the expression matrix and meta data

gse_141445 <- gse_141445[[1]]

#expression matrix

expr_141445 <- exprs(gse_141445)

#meta-data
meta_141445 <- pData(gse_141445)

#showing first five rows of the meta-data and the expression samples
head(expr_141445, n = 5)
head(meta_141445, n = 5)
#download the data into the data directory
getGEOSuppFiles("GSE141445")
```

```{r}
meta_141445
```

```{r, inlcude=FALSE, warning=FALSE}
library(glue)
dir <- "GSE141445"
filename <- "GSE141445_RAW.tar"

if (!file.exists(dir)) {
  dir.create(dir, showWarnings = FALSE)
}
# Huge Reason why I am using this instead of regular GEOquery
# the GSEXXXXX_RAW.tar that came form GEO,query only comes with
#
system(glue(
  "if [ ! -f ./{dir}/{filename} ]; then ",
  "wget -P ./{dir}/ ",
  "https://ftp.ncbi.nlm.nih.gov/geo/series/GSE141nnn/GSE141445/suppl/",
  "{filename}; ",
  "fi"
))

file.exists("./GSE141445/GSE141445_RAW.tar")


system(glue("tar -xzf ./{dir}/{filename} -C ./{dir}/"))
```

### Reading in Count Matrix, Creating Seurat Object

```{r}
library(glue)
dir <- "GSE141445"

filename <- "GSM4203181_data.raw.matrix.txt.gz"


#read in via path name from downloaded file
path <- glue("./{dir}/{filename}")
counts <- read.delim(gzfile(path))
cnt_mat <- as.matrix(counts)
rm(counts)
```

```{r}
gse141445.obj <- CreateSeuratObject(cnt_mat)
rm(cnt_mat)
```

```{r}
print("prior to editing")
gse141445.obj@meta.data
```

### Editing MetaData

**reading in clincal_data with the following variables (PatientID,
patient_num, Age, Biopsy_gleason, pre-op PSA, pT)**

```{r}
library(openxlsx)
clinical_gse141445 <- read.xlsx("./clinical_data/filtered_clinical_data_gse141445.xlsx")
rownames(clinical_gse141445) <- NULL
clinical_gse141445 <- clinical_gse141445 %>% mutate(patient_num = as.factor(patient_num),
                                                    Age = as.factor(Age),
                                                    bGleason = as.factor(bGleason),
                                                    pre_PSA = as.factor(pre_PSA),
                                                    pT = as.factor(pT))
clinical_gse141445
```

```{r}
#Adding First Author and Year
gse141445.obj@meta.data$orig.ident <- rep("Chen2021", nrow(gse141445.obj@meta.data))

#Adding GEO Accession Number
gse141445.obj@meta.data$gse_acc <- rep("GSE141445", nrow(gse141445.obj@meta.data))

#Adding Conditions for each sample:
#"Here we have profiled the transcriptomes of 36,424 single cells from 13 prostate tumours and identified the epithelial cells underlying disease aggressiveness."
#therefore each cell is from a tumor sample
gse141445.obj@meta.data$condition <- rep("tumor", nrow(gse141445.obj@meta.data))
gse141445.obj@meta.data
```

```{r}
#extracting Patient ID
#"Collected from patients who had undergone radical prostatectomy at Shanghai Changhai Hospital. The number following the barcode (the name of each column) reflects the patient number." - family.xml file in GSE141445
gse141445.obj@meta.data$patient_num <- 
  stringr::str_extract(
    rownames(gse141445.obj@meta.data),
    "[0-9]+$"
  )

gse141445.obj@meta.data <- gse141445.obj@meta.data %>%
  mutate(patient_num = as.factor(patient_num))


barcodes <- rownames(gse141445.obj@meta.data)
```

```{r}
#for some reason left_join() deletes the cell barcodes, this is why I usually don't use it
gse141445.obj@meta.data <- left_join(gse141445.obj@meta.data, clinical_gse141445, by="patient_num")

gse141445.obj@meta.data <- gse141445.obj@meta.data %>% select(-patient_num)
rownames(gse141445.obj@meta.data) <- barcodes
```

```{r}
print("after editing")
#Ideally we should have 10 columns at most
gse141445.obj@meta.data
```

### Saving GSE141445 as an RDS Object

```{r}
dir <- "sc_rds_files"
if (!file.exists(dir)) {
  dir.create(dir, showWarnings = FALSE)
}
saveRDS(gse141445.obj, glue("./{dir}/gse141445.rds"))
#removing non essential data from the

rm(gse_141445,meta_141445,expr_141445)
```

```{r}
gse141445.obj <- readRDS("./sc_rds_files/gse141445.rds")
```

```{r}
colnames(gse141445.obj@meta.data)
```

finished with pre-processing of GSE141445

Metadata contains the following columns

```         
"orig.ident"   "nCount_RNA"   "nFeature_RNA" "gse_acc"      "condition"    "PatientID"    "Age"          "bGleason"     "pre_PSA"      "pT" 
```

# GSE176031/Song, H Download

Single-cell RNA sequencing based on **Seq-Well S\^3 protocol** of:

The authors provided the Raw matrix in their GeoAccession

Confirmation from Patient IDs

- 6 Prostate biopsies from 3 different patients (Patient 1, Patient 2,
  Patient 3)

- 4 radical prostatectomies with tumor-only samples from 4 patients
  (Patients 4, 5,6, 11)

- 4 radical prostatectomies with matched normal samples from 4 patients
  (tumor-core + adj.N)

  (Patients 7,8,9,10)

- Organoids derived from radical prostatectomy paired tumor and normal
  tissue samples.

  Derived from Patients (7, 8, 9, 10, 11)

  Patients 12 and 13 are not noted within the metadata until their
  organioids progenitors are formed

\*Why are organoid necessary for prostate cancere studies?\* Answers for
later date

MetaData columns:

**MetaData Columns: (orig.ident, sample_title, disease state, patientID
(if possible))**

### Downloading Data

```{r}
library(GEOquery)
#enter in GSE141445
#read in the whole count matrices and meta data

gse_176031 <- getGEO("GSE176031", GSEMatrix = TRUE)
#divide the ExpressionSet into the expression matrix and meta data

gse_176031 <- gse_176031[[1]]

#expression matrix

expr_176031 <- exprs(gse_176031)

#meta-data
meta_176031 <- pData(gse_176031)

#showing first five rows of the meta-data and the expression samples
head(expr_176031, n = 5)
head(meta_176031, n = 5)
#download the data into the data directory
getGEOSuppFiles("GSE176031")
```

```{r}
#beginning we have 53 unique samples with unique geo_accession demarkation
meta_176031
```

```{r}
meta_176031 <- meta_176031 %>% 
  select(c(title,geo_accession, source_name_ch1,"patient:ch1", "tissue:ch1" ))
meta_176031
```

```{r}
state_markers <- c(unique(meta_176031$source_name_ch1))
marker_replacements <- c("biopsy","tumor","norm_adj","tumor_core","organoid")
state_markers
```

```{r}
#Intent is to replace the column's entries in the first character vector
# with the entries in the character vector
state_markers <- c(unique(meta_176031$source_name_ch1))
marker_replacements <- c("biopsy","tumor","norm_adj","tumor_core","organoid")

replacement <- setNames(marker_replacements, state_markers)

colnames(meta_176031)[colnames(meta_176031) == "patient:ch1"] <- "PatientID"

meta_176031 <- meta_176031 %>%
  mutate(source_name_ch1 = replacement[source_name_ch1]) %>%
  mutate(geo_source = paste0(geo_accession,"_", source_name_ch1)) %>%
  mutate(PatientID = trimws(PatientID)) %>%
  mutate(PatientID = as.factor(PatientID))
```

```{r}
meta_176031
```

Clinical Data was only available for 11 out of 13 patients in the study
including prostate organoids.

Outsdide of R I summarized a supplementary table from the original
article and imported into R to merge with the meta_data

```{r}
library(openxlsx)
clinical_data_176031 <- read.xlsx("./clinical_data/filtered_clinical_data_gse176031.xlsx")
rownames(clinical_data_176031) <- NULL
clinical_data_176031
```

```{r}
levels(meta_176031$PatientID)
```

```{r}
meta_176031 <- left_join(meta_176031, clinical_data_176031, by = "PatientID")
colnames(meta_176031)[colnames(meta_176031) == "geo_accession"] <- "sample"
#The title should not be duplicated this will balloon our sample count and our cell count down the line.
# Deduplication is necessary here because the authors extracted mulitple biopsies from a single patient and performed RNA-Seq, however they are marked as seperate samples. The clinical data I brought in labeled them with accurately with double PatientID demarcations to indicate that GSM14XXX and GSM15XXX came from Patient 1 with the same Gleason, pre-PSA (pre-operative PSA), etc. thus duplicated clinical values for different samples. However a side effect of left join is that there are multiple matches to PatientID column which not only duplicated the title, but the GSMXX number giving us redundant row
#the code below helps us to deal with that. keeping the original number of rows n = 53 while preserving the PSA, Gleason catorigal values
meta_176031 <- meta_176031[!duplicated(meta_176031$sample), ]
meta_176031
```

Somehow the Organoids have Gleason Scores. I look at the supplementary
material located in the clinical_data folder.

```{r}
meta_176031
```

Looking at the meta data available from GEOquery API call and the
Supplmentary Data that carries the clinical data. It looks like the only
way to marry the two is the PatientID column.

```{r, include=FALSE}
library(glue)

dir <- "GSE176031"
ftp <- "https://ftp.ncbi.nlm.nih.gov/geo/series/GSE176nnn/GSE176031/matrix/"
filename <- "GSE176031_RAW.tar"

if (!file.exists(dir)) {
  dir.create(dir, showWarnings = FALSE)
}
# Huge Reason why I am using this instead of regular GEOquery
# the GSEXXXXX_RAW.tar that came form GEO,query only comes with
#
system(glue(
  "if [ ! -f ./{dir}/{filename} ]; then ",
  "wget -P ./{dir}/ ",
  "{ftp}",
  "{filename}; ",
  "fi"
))

system(glue("tar -xzf ./{dir}/{filename} -C ./{dir}/"))
```

```{r}

#We don't need the organiod samples.
# 1 they are not within out perview of research. It answers a totally different question
# 2 It will save on data and I may be able to use more healthy samples instead of organoid


files <- list.files(path = "./GSE176031",
                    pattern = "\\_dge.txt.gz",
                    recursive = TRUE,
                    full.names = TRUE)

regex <- "org|org_pool|_org_"
files <- files[!grepl(regex, files)]
files
```

I am pre-processing a dataset form GEOAccession The final feature space
of GSE176031 when collumn bound has 9007 genes. In any highthroughput
assay that is really small. When integrating that with other datasets
down the line, this will hurt the analysis. When reading in the list of
samples, there is a range of 14,000 genes to 23,000 genes. How would I
keep as many genes possible?

```{r}

cnt_mats_176031 <- list()
for(i in seq_along(files)){
  cnt_mats_176031[[i]] <- read.delim(gzfile(files[i]))
  
  cnt_mats_176031[[i]] <- column_to_rownames(
    cnt_mats_176031[[i]],
    "GENE")
  
  #cat("number of genes:",dim(cnt_mats_176031[[i]])[1], "number of cells:",dim(cnt_mats_176031[[i]])[2], "\n")
}
names(cnt_mats_176031) <- sub("^(GSM[0-9]+_).*", "\\1", basename(files))
names(cnt_mats_176031) <- sub("_", "", names(cnt_mats_176031))


#currently the GeoQuery Metadata has all of the samples it needs to reflect that ther has been changes to the sample list
meta_176031 <- meta_176031 %>% filter(sample %in% names(cnt_mats_176031))
names(cnt_mats_176031) <- unique(meta_176031$geo_source)
```

### Barcode Editing

Not all of the Barcodes are unique on a global scale of the whole
dataset.

The code below will identify non-unique barcodes and add a .i+1
demarcation. Then will harmonize the genes asayed for each sample to
column bind them into one matrix.

```{r}
#iF WE DO UNION THEN THAT WILL BALLOON THE SIZE OF sEURAT oBJECT TO
# 6.5GB PRRIOR TO DIM REDUCTION
common_features_176031 <- Reduce(intersect, lapply(cnt_mats_176031, rownames))
length(common_features_176031)

cnt_mats_common_176031 <- lapply(
  cnt_mats_176031,
  \(x) x[common_features_176031, ,drop = FALSE]
)
# for some reason this is appending the list names onto the colnames
 # Which give global uniqueness to the barcodes
#However take that away, and we have non-unique barcodes between sequencing libraries
#I need to figure out why the names(cnt_mats) gets pre-pended to the DNA barcodes
count_matrix_176031 <- do.call(cbind, cnt_mats_common_176031)
#rm(cnt_mats)
#rm(cnt_mats_common)
length(unique(rownames(count_matrix_176031))) == nrow(count_matrix_176031)
dim(count_matrix_176031)
rm(cnt_mats_common_176031, cnt_mats_176031)
```

```{r}
gc()
```

Checking globalm uniquenes not inter-library uniqueness

```{r}
#rm(cnt_mats)
#rm(cnt_mats_common)
count_matrix_176031 <- as.matrix(count_matrix_176031)
length(unique(rownames(count_matrix_176031))) == nrow(count_matrix_176031)

#There is global unique-ness between the barcodes, but not uniqueness between libraries
anyDuplicated(colnames(count_matrix_176031))
any(is.na(colnames(count_matrix_176031)))
any(colnames(count_matrix_176031) == "")
```

Editing Rownames before reading into the Seurat

```{r}
#you can't edit Cell Names within the assay after creating the Seurat object
assay_colnames <- colnames(count_matrix_176031)

#seperating the cell_names from the GSMXXXXX sample title
sample_title <- sub("\\..*", "", assay_colnames)

# this will address the inter library non-unique scenario
non_unique_barcodes <- sub("^[^.]*\\.", "", assay_colnames)

n <- ave( non_unique_barcodes, non_unique_barcodes, FUN = seq_along)

result <- paste0(non_unique_barcodes, ".", n)

colnames(count_matrix_176031) <- result
```

### Editing Seurat Meta-Data

```{r}
gse176031.obj <- CreateSeuratObject(count_matrix_176031)
rm(count_matrix_176031)
gse176031.obj@meta.data$orig.ident <- "Song2021"
gse176031.obj@meta.data$geo_acc <- "GSE176031"

gse176031.obj@meta.data$sample_title  <- sample_title

#Barcode IDs are changed in the meta.data
#but we need to address the data within Assay as well
gse176031.obj@meta.data <- gse176031.obj@meta.data %>%
  mutate(condition = str_extract(sample_title, "(?<=_).*")) %>%
  mutate(sample = str_extract(sample_title, "^[^_]+")) %>%
  select(-sample_title)
```

```{r}
gse176031.obj@meta.data <- left_join(gse176031.obj@meta.data, meta_176031, by = "sample")
```

```{r}
DefaultAssay(gse176031.obj) <- "RNA"
length(colnames(gse176031.obj)) == length(result)
rownames(gse176031.obj@meta.data) <- result
```

```{r}
saveRDS(gse176031.obj, "./sc_rds_files/gse176031.rds")
```

```{r}
rm(cnt_mats_176031, 
   cnt_mats_common_176031, 
   count_matrix_176031, 
   meta_176031, 
   gse_176031,
   gse176031.obj)
```

Problem: The authors stated that they had \~21,000 cells after QC.
However after concatenating we have an insane 53,00 cells. It's very
likely these are not QCed but it's a concerning amount to be that far
from the Paper's numbers

We'll have to see after QC if this reduces at all highly likely it will
reduce these cells by alot

# [GSE150692 (Crowley, 2020) (PRAD](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE150692)) Download

2 normal benign prostate tissue specimens obtained from patient
resections.

**10x Genomics Chromium platform**

In the Supplementary Data Crowley, et.al state that they used 6 human
prostate samples, but only 3 are availble for download in the GEOquery
website. BRUH

**Human Samples Available on GEOQuery:**

GSM4556600(MF002) clinical data available, benign tumor

GSM4556601(MM033), benign tumor

GSM4556602(MM037) , benign tumor

**Mouse Samples (GeoQuery MetaData only gives the GSMXXXs for mouse
Samples):**

GSM4556594 GSM4556595 GSM4556596 GSM4556597 GSM4556598 GSM4556599

DO NOT INCLUDE MOUSE SAMPLES LOL

```{r}
library(GEOquery)
#enter in GSE141445
#read in the whole count matrices and meta data

gse_150692 <- getGEO("GSE150692", GSEMatrix = TRUE)
#divide the ExpressionSet into the expression matrix and meta data

gse_150692 <- gse_150692[[1]]

#expression matrix

expr_150692 <- exprs(gse_150692)

#meta-data
meta_150692 <- pData(gse_150692)

#showing first five rows of the meta-data and the expression samples
head(expr_150692, n = 5)
head(meta_150692, n = 5)
#download the data into the data directory
getGEOSuppFiles("GSE150692")
```

```{r}
meta_150692
```

### GEOquery Pull Excludes Human Assays

```{r}
library(glue)
dir <- "GSE150692"
filename <- "GSE150692_RAW.tar"

if (!file.exists(dir)) {
  dir.create(dir, showWarnings = FALSE)
}
# Huge Reason why I am using this instead of regular GEOquery
# the GSEXXXXX_RAW.tar that came form GEO,query only comes with
#
system(glue(
  "if [ ! -f ./{dir}/{filename} ]; then ",
  "wget -P ./{dir}/ ",
  "https://ftp.ncbi.nlm.nih.gov/geo/series/GSE150nnn/GSE150692/suppl/",
  "{filename}; ",
  "fi"
))

system(glue("tar -xzf ./{dir}/{filename} -C ./{dir}/"))

#GSMXX60 Are all human samples, but one of the sample files was not named with human marker

system(glue(
  "mv ./{dir}/GSM4556602_MM037_clean_raw_counts_matrix.txt.gz ",
  "./{dir}/GSM4556602_MM037_human_clean_raw_counts_matrix.txt.gz"
))

#sweep for only human files in the downloaded data
files <- list.files(path = "./GSE150692",
                    pattern = "_human_.*\\.txt\\.gz$",
                    recursive = TRUE,
                    full.names = TRUE)
                    
basename(files)

cnt_mats_150692 <- list()
for(i in seq_along(files)){
  cnt_mats_150692[[i]] <- read.delim(gzfile(files[i]))
}



names(cnt_mats_150692) <- sub("^(GSM[0-9]+_).*", "\\1", basename(files))
names(cnt_mats_150692) <- sub("_", "", names(cnt_mats_150692))
```

```{r}
#finding repeat features within the same matrix
common_features <- Reduce(intersect, lapply(cnt_mats_150692, function(x) x$X))

common_features[1:20]
```

### Fixing Repeated Gene Names in the data (Warning, Very Long Wait)

```{r}
#condensing only repeated features per column this is not pseudobulking
cnt_mats_common_150692 <- lapply(cnt_mats_150692, function(x) {
  x %>%
    group_by(X) %>%
    summarise(across(where(is.numeric), sum)) %>%
    filter(X %in% common_features)
})
# for some reason this is appending the list names onto the colnames
# Which give global uniqueness to the barcodes
#However take that away, and we have non-unique barcodes between sequencing libraries
#I need to figure out why the names(cnt_mats) gets pre-pended to the DNA barcodes
count_matrix_150692 <- do.call(cbind, cnt_mats_common_150692)
```

```{r}
count_matrix_150692 <- as.tibble(count_matrix_150692) %>% column_to_rownames("GSM4556600.X")
```

```{r}
rm(cnt_mats_150692)
rm(cnt_mats_common_150692)
count_matrix <- as.matrix(count_matrix_150692)
dim(count_matrix)
length(unique(rownames(count_matrix_150692))) == nrow(count_matrix_150692)

#There is global unique-ness between the barcodes, but not uniqueness between libraries
anyDuplicated(colnames(count_matrix_150692))
any(is.na(colnames(count_matrix_150692)))
any(colnames(count_matrix_150692) == "")
```

```{r}
gse150692.obj <- CreateSeuratObject(count_matrix_150692)
```

Editing Meta Data

Human samples only from GSE150692

```{r}
gse150692.obj

gse150692.obj@meta.data$orig.ident <- "Crowley2020"
gse150692.obj@meta.data$geo_acc <- "GSE150692"

cell_names <-  rownames(gse150692.obj@meta.data)

sample_title <- sub("\\..*", "", cell_names)

gse150692.obj@meta.data$sample_title  <- sample_title

# this will address the inter library non-unique scenario
# this will address the inter library non-unique scenario
non_unique_barcodes <- sub("^[^.]*\\.", "", cell_names)

n <- ave(non_unique_barcodes, non_unique_barcodes, FUN = seq_along)

result <- paste0(non_unique_barcodes, ".", n)

rownames(gse150692.obj@meta.data) <- result

```

```{r}
gse150692.obj@meta.data
saveRDS(gse150692.obj, "./sc_rds_files/gse150692.rds")
```

```{r}
gse150692.obj <- readRDS("./sc_rds_files/gse150692.rds")

```

```{r}
dim(gse150692.obj@assays$RNA)
```

# GSE181294 ([Chen et. al., 2023](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE181294))

Sequenced with Illumina NextSeq 500, Prepped with 10x Chromium Protocol
(Single Cell 3' Library and Gel Bead Kit v2)

DO NOT DOWNLOAD THE SPATIAL DATA

Hold on in the paper they mentioned that they used sc as well as spatial
data. How do I know which is Spatial Data and which is Single cell data.
These hints are actually given in the annotation files listed as

**GSE181294_slide.seq.ano.csv and GSE181294_scRNAseq.ano.csv**

**Spatial Data Samples: S1\_**HP1, S2_HP2, S3_HP3, S4_HP4, S5_Benign01,
S6_Benign02, S7_Benign03, S8_Benign04, S9_Tumor01, S10_Tumor02,
S11_Tumor03, S_11Tumor04. These are also all of the **S[0-9]**
demarcated samples. Which is why they have BARCODES and GENES as
seperate csv vecor SO DO NOT DOWNLOAD THOSE LOL

**Single Cell Data Samples: Essentially anything without the prefix
\^S[0-9] or regex pattern \^GSM61339[0-9]+\\\\.counts\\\\.mtx\\\\.gz**

**20 human prostate tumors + matched adj Normal tissues**

**5 Additional Healthy Prostates**

It may be waste of time to go after all of the files in this study
instead I may just opt in to grab the Healthy Data. This will make
downloading this data more suitable for both the Windows PC and Mac
book.

```{r}
library(GEOquery)
#enter in GSE141445
#read in the whole count matrices and meta data

gse_181294 <- getGEO("GSE181294", GSEMatrix = TRUE)
#divide the ExpressionSet into the expression matrix and meta data

gse_181294 <- gse_181294[[1]]

#expression matrix

expr_181294 <- exprs(gse_181294)

#meta-data
meta_181294 <- pData(gse_181294)

#showing first five rows of the meta-data and the expression samples
head(expr_181294, n = 5)
head(meta_181294, n = 5)
#download the data into the data directory
getGEOSuppFiles("GSE181294")
```

```{r}
meta_181294
```

### Geoquery Meta-Data

```{r}
colnames(meta_181294)
meta_181294 <- meta_181294 %>% 
  select(c(title, geo_accession, "condition:ch1", "gleason:ch1")) %>%
  mutate(`gleason:ch1` = case_when(
                                  "NA" == `gleason:ch1` ~ NA_character_,
                                  str_detect(`gleason:ch1`, "=") ~ gsub("=.*$", "", `gleason:ch1`))) %>%
          mutate(`gleason:ch1` = trimws(`gleason:ch1`))

#before renaming
meta_181294

#renaming columns
colnames(meta_181294)[colnames(meta_181294) == "gleason:ch1"] <- "bGleason"
colnames(meta_181294)[colnames(meta_181294) == "condition:ch1"] <- "condition"

meta_181294
```

```{r}
condition_markers <- c(unique(meta_181294$condition))
condition_label_replacements <- c("healthy", "norm_adj", "tumor_core")

replacement <- setNames(condition_label_replacements, condition_markers)

meta_181294 <- meta_181294  %>%
  mutate(condition = replacement[condition])
```

### Reading in Count Matrices for GSE181294

```{r}
library(glue)
dir <- "GSE181294"
filename <- "GSE181294_RAW.tar"

if (!file.exists(dir)) {
  dir.create(dir, showWarnings = FALSE)
}
# Huge Reason why I am using this instead of regular GEOquery
# the GSEXXXXX_RAW.tar that came form GEO,query only comes with
#
system(glue(
  "if [ ! -f ./{dir}/{filename} ]; then ",
  "wget -P ./{dir}/ ",
  "https://ftp.ncbi.nlm.nih.gov/geo/series/GSE181nnn/GSE181294/suppl/",
  "{filename}; ",
  "fi"
))

system(glue("tar -xzf ./{dir}/{filename} -C ./{dir}/"))
```

There are two types of files that exist in GSE181294: completed count
matrices and fragmented count matrices where the barcodes and genes are
stored in vectors. It's likely the author uploaded this way due to some
kind of memory guidelines. Which makes sense for 51 seperate count
matrices with likely over 100,000 cells available. We have to be very
VERY conscious on how to handle this read in.

In order not to fry my new RAM sticks or my MOBO, I'll have to parse out
the chunks of data carefully. Luckily it served the first run through
with a more efficient loop but something tells me I'll be coming back to
fix some metadat and I'll have to reprocess some stuff.

```{r}
library(Matrix)
library(data.table)
#system statistics
library(ps)
#extracting files list in the GEOquery metadata
#CSV count files
# Matrices are too big for my local computer. It's time to compromise
# Only use the Health cells as a reference
regex2 <- ".*\\_HP[0-9]+\\.count\\.csv\\.gz$"
#matching to only 10 pairs of tumor_core/adj_norm datasets
regex1 <- "^GSM54943(4[7-9]|5[0-9]|6[0-6])\\_SCG.*\\.count\\.csv\\.gz$"
patterns_count_files <- c(regex1,regex2)

count_files <- list.files(path = "./GSE181294",
                          pattern = paste0(patterns_count_files, collapse = "|"),
                          recursive = TRUE,
                          full.names = TRUE)

basename(count_files)
```

```{r}
cnt_mats_181294 <- list()
read_count_matrix <- function(f) {
    if (grepl("\\.csv(\\.gz)?$", f)) {
    data.table::fread(f)
    x <- data.frame(data.table::fread(f, check.names = FALSE))
    rownames(x) <- x[[1]]
    x[[1]] <- NULL
  } else {
    stop("Unsupported file type: ", f)
  }
return(x)}

cnt_mats_181294 <- lapply(count_files, read_count_matrix)

names(cnt_mats_181294) <- sub("^(GSM[0-9]+_).*", "\\1", basename(count_files))
names(cnt_mats_181294) <- sub("_", "", names(cnt_mats_181294))
```

```{r}
gc()
cnt_mats_181294$GSM5494343
```

```{r}
common_features_181294 <- Reduce(intersect, lapply(cnt_mats_181294, rownames))


cnt_mats_common_181294 <- lapply(
  cnt_mats_181294,
  \(x) x[common_features_181294, ,drop = FALSE]
)
# for some reason this is appending the list names onto the colnames
# Which give global uniqueness to the barcodes
#However take that away, and we have non-unique barcodes between sequencing libraries
#I need to figure out why the names(cnt_mats) gets pre-pended to the DNA barcodes
count_matrix_181294 <- do.call(cbind, cnt_mats_common_181294)
#rm(cnt_mats)
#rm(cnt_mats_common)
#checking uniqueness on the rows
length(unique(rownames(count_matrix_181294))) == nrow(count_matrix_181294)
```

```{r}
rm(cnt_mats_181294)
rm(cnt_mats_common_181294)
count_matrix_181294 <- as.matrix(count_matrix_181294)
length(unique(rownames(count_matrix_181294))) == nrow(count_matrix_181294)

#checking for any duplicates on the colnames
anyDuplicated(colnames(count_matrix_181294))
any(is.na(colnames(count_matrix_181294)))
any(colnames(count_matrix_181294) == "")
```

### Creating Seurat Object

I will add to the meta data on a different computer

```{r}
rm(count_matrix_176031)
gse181294.obj <- CreateSeuratObject(count_matrix_181294)
gse181294.obj@meta.data$orig.ident <- "Chen2023"
gse181294.obj@meta.data$geo_acc <- "GSE181294"
```

```{r}
rm(count_matrix_181294)
saveRDS(gse181294.obj, "./sc_rds_files/gse181294.rds")
```

MetaData - ClinicalData Merging

```{r}
gse181294.obj <- readRDS("./sc_rds_files/gse181294.rds")
```

```{r}
gse181294.obj@meta.data

#extractig the cell names
cell_names <-  rownames(gse181294.obj@meta.data)
cell_names[1:10]
#seperating the cell_names from the GSMXXXXX sample title
#sample_title <- sub("\\..*", "", cell_names)
barcodes <- sub("^[^.]*\\.","", cell_names)
barcodes[1:10]
sample_title <- str_extract(barcodes, "^[^_]*_") %>% str_remove("_")

if(!identical(barcodes, rownames(gse181294.obj@meta.data))){ 
  rownames(gse181294.obj@meta.data) <- barcodes 
  }

gse181294.obj@meta.data$title <- sample_title

gse181294.obj@meta.data <- left_join(gse181294.obj@meta.data, meta_181294, by = "title")
```

```{r}
gse181294.obj@meta.data
if(!identical(barcodes, rownames(gse181294.obj@meta.data))){ 
  rownames(gse181294.obj@meta.data) <- barcodes 
}
gse181294.obj@meta.data
```

```{r}
saveRDS(gse181294.obj, "./sc_rds_files/gse181294.rds")
```

**Stop Here I think 4 is enough lol. The pursuit of excellence is noble,
but the pursuit of perfection is neurotic**

Next Step: Write Sample Chart for the collected scRNA-Seq samples

# Final Sample Sheet

+-----------+----------------------------------------------------------------------+-------------+--------------+----------+----------+---------------+--------------+
|           | Author(s)                                                            | Organism(s) | Condition(s) | preQC    | Lib Prep | Availability  | Clinical     |
|           |                                                                      |             |              | Reported | Protocol | Format        | Data         |
|           |                                                                      |             |              | Cell Cnt |          |               |              |
+===========+======================================================================+=============+==============+==========+==========+===============+==============+
| GSE141445 | [(Chen et. al.,                                                      | Human       | Tumor        | 36,424   | 10x      | series_matrix | available,   |
|           | 2021)](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE141445) |             |              |          | Chromium |               | but sparse   |
|           |                                                                      |             |              |          | V2       |               |              |
+-----------+----------------------------------------------------------------------+-------------+--------------+----------+----------+---------------+--------------+
| GSE176031 | [(Song, H                                                            | Human       | Paired       | 53,000   | Seq-Well | series_matrix | available,   |
|           | 2021)](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE176031) |             | (AdjNorm +   |          |          |               | but sparse   |
|           |                                                                      |             | tumor_core)  |          |          |               | and          |
|           |                                                                      |             |              |          |          |               | inconsistent |
+-----------+----------------------------------------------------------------------+-------------+--------------+----------+----------+---------------+--------------+
| GSE150692 | [(Crowley,                                                           | Human / Mus | Tumor        |          | Chromium | series_matrix | sparse but   |
|           | 2020)](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE150692) | Musculus    |              |          | Single   |               | available    |
|           |                                                                      |             |              |          | Cell 5’  |               |              |
|           |                                                                      | "orthotopic |              |          | Kit      |               | only         |
|           |                                                                      | 6694c2      |              |          |          |               |              |
|           |                                                                      | pancreatic  |              |          | DO NOT   |               | 1 out of     |
|           |                                                                      | tumors"     |              |          | USE      |               | three has    |
|           |                                                                      |             |              |          |          |               | useful       |
|           |                                                                      |             |              |          |          |               | clinical     |
|           |                                                                      |             |              |          |          |               | data.        |
|           |                                                                      |             |              |          |          |               |              |
|           |                                                                      |             |              |          |          |               | Not worth    |
|           |                                                                      |             |              |          |          |               | trying to    |
|           |                                                                      |             |              |          |          |               | use the      |
|           |                                                                      |             |              |          |          |               | clinical     |
|           |                                                                      |             |              |          |          |               | data         |
+-----------+----------------------------------------------------------------------+-------------+--------------+----------+----------+---------------+--------------+
| GSE181294 | [(](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi)[Chen et. al.,    | Human       | Healthy      |          | 10x      | series matrix | only Gleason |
|           | 2023)](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE181294) |             |              |          | Chromium |               | Available    |
|           |                                                                      |             | and Paired   |          | V2       | only using 10 |              |
|           |                                                                      |             | (PRAD / N.   |          |          | paired        |              |
|           |                                                                      |             | Adj)         |          |          | samples and 5 |              |
|           |                                                                      |             |              |          |          | healthy ones  |              |
+-----------+----------------------------------------------------------------------+-------------+--------------+----------+----------+---------------+--------------+

# GSE120716 [(Henry et. al., 2018)](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE120716)

GSE1207816 is a SuperSeries of two different resolutions of RNA-Seq data

[GSE117271](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE117271):
Bulk RNAseq data of cell types isolated by FACS from normal human
prostates

[GSE117403](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE117403):
scRNA-Seq of cell types isolated by FACS from normal human prostates

```{r}
library(GEOquery)
#enter in GSE141445
#read in the whole count matrices and meta data

gse_117403 <- getGEO("GSE117403", GSEMatrix = TRUE)
#divide the ExpressionSet into the expression matrix and meta data

gse_117403 <- gse_117403[[1]]

#expression matrix

expr_117403 <- exprs(gse_117403)

#meta-data
meta_117403 <- pData(gse_117403)

#showing first five rows of the meta-data and the expression samples
head(expr_117403, n = 5)
head(meta_117403, n = 5)
#download the data into the data directory
getGEOSuppFiles("GSE117403")
```

```{r}
meta_117403
```

```{r}
dir <- "GSE117403"
count_files <- list.files(path = "./GSE117403",
                          pattern = "^GSE117403.*",
                          recursive = TRUE,
                          full.names = TRUE)
#save the names of the files to a character vector
count_files <- basename(count_files)

D17 <- str_extract(count_files, "^GSE117403_D17_.*")
D17 <- D17[!is.na(D17)]
D27 <- str_extract(count_files, "^GSE117403_D27_.*")
D27 <- D27[!is.na(D27)]
Pd <- str_extract(count_files, "^GSE117403_Pd_.*")
Pd <- Pd[!is.na(Pd)]

count_files <- list(
  D17 = D17,
  D27 = D27,
  Pd  = Pd
)
count_files
cnt_mat_list_117403 <- list()

for (i in names(count_files)) {
  
  files <- count_files[[i]]
  
  mtx_file <- files[str_detect(files, "\\.mtx\\.gz$")]
  features_file <- files[str_detect(files, "features|genes")]
  barcodes_file <- files[str_detect(files, "barcodes")]
  
  cnt_mat <- Matrix::readMM(gzfile(glue("./{dir}/{mtx_file}")))
  
  features <- read_tsv(gzfile(glue("./{dir}/{features_file}")), col_names = FALSE)
  rownames(cnt_mat) <- features[[1]]
  
  barcodes <- read_tsv(gzfile(glue("./{dir}/{barcodes_file}")), col_names = FALSE)
  colnames(cnt_mat) <- barcodes[[1]]
  
  cnt_mat_list_117403[[i]] <- cnt_mat
  
}
```

```{r}
common_features_117403 <- Reduce(intersect, lapply(cnt_mat_list_117403, rownames))


cnt_mats_common_117403 <- lapply(
  cnt_mat_list_117403,
  \(x) x[common_features_117403, ,drop = FALSE]
)
# for some reason this is appending the list names onto the colnames
# Which give global uniqueness to the barcodes
#However take that away, and we have non-unique barcodes between sequencing libraries
#I need to figure out why the names(cnt_mats) gets pre-pended to the DNA barcodes
count_matrix_117403 <- do.call(cbind, cnt_mats_common_117403)
#rm(cnt_mats)
#rm(cnt_mats_common)
#checking uniqueness on the rows
length(unique(rownames(count_matrix_117403))) == nrow(count_matrix_117403)
```
