---
title: "Untitled"
output: html_document
---

```{r}
library(Seurat)
library(tidyverse)
library(ggplot2)
library(hdf5r)
library(harmony)
#library(BPCells)
library(SeuratWrappers)
library(GEOquery)
```

```{r}
gse181294.obj <- readRDS("./post_qc_rds_files/gse181294.rds")
gse150692.obj <- readRDS("./post_qc_rds_files/gse150692.rds")
gse176031.obj <- readRDS("./post_qc_rds_files/gse176031.rds")
gse141445.obj <- readRDS("./post_qc_rds_files/gse141445.rds")
```

```{r}
sum(nrow(gse181294.obj@meta.data), nrow(gse150692.obj@meta.data), nrow(gse176031.obj@meta.data), nrow(gse141445.obj@meta.data))
```

Meta Data Correction Check list

At minimum I need the following columns: orig.ident: the author geo_acc: the GSEXXXXX number condition: biopsy,healthy,tumor_core,normal_adj

Fixes check list:

**GSE181294**

Current meta data columns:

**orig.ident, gse_acc**

- does not have a condition column

- need to move the sample ID over to a column (sample_ID)

**GSE150692**

orig.ident, gse_acc

- need to add the GSM/Sample ID

- need to add condition column

**GSE176031**

- column geo_source is redundant, delete it

**GSE141445**

- add sample ID GSM number

```{r}
gse181294.obj@meta.data
gse150692.obj@meta.data
gse176031.obj@meta.data
gse141445.obj@meta.data
```
