---
title: "SC_QualityControl"
output: html_document
---

# Library Read In

```{r}
#Seurat Core
library(Seurat)
#library(BPCells)
library(SeuratWrappers)
library(S4Vectors)

#non-Saitja QC libraries
library(DropletUtils)
library(scDblFinder)
library(SoupX)
library(SingleCellExperiment)

#Visualizations
library(ggplot2)
library(patchwork)
library(dplyr)
library(tidyverse)
library(glue)

#utility function
source("utility_scripts.R")
```

nfeature: represents the total number of unique genes detected in each individual cell

nCount: total sum of all UMI counts detected across all genes within that cell

# QC Strategy

### Do the datasets even NEED QC?

For example **GSE141445'**s paper stated that "After standard data processing and quality control procedures (Methods), we obtained transcriptomic profiles for 36,424 cells (Extended Data Fig. 1a)". This portion of cells was uploaded to GEO Accession website so therefore it's unnecessary for further QC. Downsampling further on this dataset could either dampen relevant signals or cut out weaker signals entirely. –\> No

**GSE176031** does need the data processed –\> "Raw single-cell RNA-sequencing FASTQ files and gene expression matrices files generated in this study have been deposited in the Gene Expression Omnibus (GEO) under accession number [GSE176031](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE176031)" QC is necessary for this even with the cut organoids –\> **Yes check it**

**GSE150692** Although the files say they are clean, it's not clear in the data availability section of the paper. It's work checking –\> **Yes**

**GSE181294:** Data availability states that "processed data can be accessed from NCBI Gene Expression Omnibus datase" however it's not stated that QCed data is available. However it's likely that emptydroplets have been taken care of. –\> Check

Resort Automatic Strategies first you've completed the manual Find EmptyDroplets (DropletUtils) Find AmbientRNA (SoupX) Find Doublets (DoubletFinder).

**Manual QC comes after automated Processing when plotting Violin plots and scatterplots look for Bimodal distributions Outliers in scatter plot Reltionships between metrics**

### **Hierarchy of thresholds**

nFeature_RNA (set threshold) to catch doublets or multiplets above the high threshold and empty droplets on the lower threshold

*nfeature_min / nfeature_max*: set thresholds for the set of cells that contain above a certain number of feature per recorded library (barcode).The purpose of he lower part will exclude potential empty or ambient RNA droplets While the maximum threshold's purpose is to exclude potenital doublets with a lot more unnecessart and possible assessment changing information of UMIs within 1 droplet but is recorded on 1 barcode.

Common thresholds(Check the visuals always)

nfeature_max = 5000

nfeature_min = 500

*ncount_min / ncount_max*: thresholds for the total UMIs within 1 column to determine what a co-encapsulated cell look within the data. This will set threshold at the nCount_RNA data

Common thresholds

ncount_min = 800

ncount_max = 20,000

nCount_RNA vs nFeature_RNA: Helps us to determine the complexity of a cell library *UMI vs Genes Detected (nCount_RNA vs nFeature_RNA)* Poor quality cells are likely to have low genes and UMIs per cell. Good cells/droplets exhbit both higher number of genes per cell and higher numbers of UMIs.

Good Combinations high UMI + high Feature Cnt UMI = high Feature Cnt Bad Combinatons extremely high UMI + extremely high Feature Cnt

*mt-thresh*: Determines what health level the "cells" should be within the droplet. Sometime cells could be in the process of dying out and thus the barcodes will capture mRNA from mitochondria. Ideal is \<5% - 10%. However this is dependent on the purpose of the study. If you're studying DYING CELLs keep it high.

ribosomal-RNA%: High ribosomal percentage is not always "bad". I could indicate activity or specific cell types like plasma stem cells, but trasking it help to detect technical batch effect or hyper-transscribing cell states

Novelty Score ( log10(nFeature)/ log10(nCout))

Captures the ratio of unique genes detected relative to total sequencing depth. Good quality cells generally have values \> 0.80 . Values below $0.70$–$0.75$ indicate low-complexity libraries (e.g., a few overrepresented transcripts, highly degraded samples, or high contamination).

```{r}
overall_dir <- "qc_images"

if(!dir.exists(overall_dir)){
  dir.create(glue("./{overall_dir}"))
}
```

# GSE141445/Chen, O QC

```{r}
gse141445.obj <- readRDS("./sc_rds_files/gse141445.rds")
```

```{r}
colnames(gse141445.obj)[1:10]
```

Estimate Sparsity

```{r}
object <- gse141445.obj
```

```{r}
#Calculate Matrix Sparsity
total_element_wise_counts <- (sum(object@assays$RNA@layers$counts) > 0)
dims <- (nrow(object) * ncol(object))
sparsity <- 1 - (total_element_wise_counts / dims)

cat("Matrix sparsity:", round(sparsity * 100, 1), "%\n")

#UMI distribution - bimodal obviously
umi_counts <- colSums(object@assays$RNA@layers$counts)
cat("Median UMI per droplet:", median(umi_counts), "\n")
cat("Droplets with >500 UMI:", sum(umi_counts > 500) , 
    "(likely cells, will be validated by EmptyDrops)\n")
```

### Empty Droplet Detection

Need to write write description of the empty Droplets functions.

```{r}
empty_results <- emptyDrops_wrapr(object@assays$RNA@layers$counts, 
                                  object,
                                  lower = 100, 
                                  niters = 10000)
```

```{r}
#Matri already filtered by CellRanger
empty_results <- object@assays$RNA$counts
```

### Setup Channel for SoupX

Preparing Soup Channel

```{r}
#Prepare SoupX channel
#tod (table of droplets) = raw counts from ALL droplets
#toc (table of cells) = filtered counts from validation cells only
sc <- SoupChannel(tod = object@assays$RNA$counts, 
                  toc = empty_results)
rownames(sc$metaData) <- colnames(object)
```

Ambient RNA checking

Setting up Temporary Object

```{r}
#basic dimension reductions not set in stone because these are important
#For cluster annotation
object <- NormalizeData(object, verbose=F)
object <- FindVariableFeatures(object, nfeatures = 5000, verbose=F)
object <- ScaleData(object, verbose = FALSE)
object <- RunPCA(object, npcs=15, verbose = FALSE)
object <- FindNeighbors(object, dims = 1:15, verbose=F)
object <- FindClusters(object, resolution = 0.8, verbose =F)
object <- RunUMAP(object, dims = 1:15, verbose=F)
```

```{r}
barcode_cluster <- setNames(as.character(object$seurat_clusters), colnames(object))
#setting clusters in the meta-data object
sc <- setClusters(sc, barcode_cluster)
```

Estimating Contanimation

```{r}
sc <- tryCatch({
  autoEstCont(sc, verbose = FALSE)
}, error = function(e) {
  cat("Note: autoEstCont failed\n")
  sc$fit$rho <- NULL
  return(sc)
})

contamination_fraction <- sc$fit$rho

# Check contamination level
cat("Contamination fraction:", 
    ifelse(is.null(contamination_fraction), "NULL (very clean sample)", 
           paste0(round(contamination_fraction * 100, 2), "%")), "\n")
 
# Decision: Skip or correct based on contamination level
if (is.null(contamination_fraction) || contamination_fraction < 0.05) {
  cat("→ Contamination is NULL or <5% - skipping SoupX correction\n")
  cat("   Your sample is clean! Proceeding with original counts.\n")
} else {
  cat("→ Contamination is", round(contamination_fraction * 100, 2), "% (≥5%)\n")
  cat("   Applying SoupX correction...\n")
 
  # Apply correction
  suppressWarnings({
    corrected_counts <- adjustCounts(sc)
  })
  object <- SetAssayData(object, layer = "counts", new.data = corrected_counts)
  cat("   ✓ SoupX correction applied\n")
}
```

Doublet Detection Doublets are when two cell or nucei are contained within a droplet after partition instead of the intended co-encapsulation 10x CellRanger captures \~1% doublets per 1000 cell load When you have a higher loading you'll likely have more doublets (10,000 cells \~6% - 9% doublets)

Rates \>10% suggest problems with cell loading concentration or the doublet detector being too sensitive. Given that our example sample includes over 9,800 cells, a 10% rate is considered acceptable.

**This can create false "hybrid" cell types which can**

1\. Inflate Cell counts

2\. Confound trajectory analysis

3\. Mess up differential expression

```{r}
#as.SingleCellExperiment expect one layer counts before adding
object <- JoinLayers(object)

sce <- as.SingleCellExperiment(object)

set.seed(42)
suppressWarnings({
  sce <- scDblFinder(sce, 
                     dbr = NULL, 
                     verbose = FALSE)
})

# Extract results
object$doublet_class <- ifelse(sce$scDblFinder.class == "doublet", "Doublet", "Singlet")
object$doublet_score <- sce$scDblFinder.score

#Tracking cells before doublet removal for QC summary
n_doublets <- sum(object$doublet_class == "Doublet")
doublet_rate <- n_doublets / ncol(object) * 100
cat("Doublets:", n_doublets, "(", round(doublet_rate, 1), "%)\n")

cells_before_doublet_removal <- ncol(object)
cat("Cells before doouble removal:", cells_before_doublet_removal)

p3 <- DimPlot(
  object,
  reduction = "umap",
  group.by = "doublet_class",
  cols = c("Singlet" = "#90E0EF", "Doublet" = "#EF233C"),
  pt.size = 1
) +
  labs(title = paste0("Doublet Detection (", round(doublet_rate, 1), "%)")) +
  theme(plot.title = element_text(face = "bold", size = 14))
```

```{r}
p3
```

```{r}
ggsave("qc_images/02_doublets_umap.png", p3, width = 8, height = 7, dpi = 300)

object <- subset(object, subset = doublet_class == "Singlet")

cat("After doublet removal:", ncol(object), "cells\n")
```

### manual thresholding for Cell level quality

```{r}
# Calculate QC metrics

#percentage of mitochondrial RNA counts
#calculate percentageof mitochondrial RNA per barcode
gse141445.obj[["percent.mt"]] <- PercentageFeatureSet(gse141445.obj, pattern = "^MT-")
#percentage of ribosomal RNA counts
gse141445.obj[["percent.ribo"]] <- PercentageFeatureSet(gse141445.obj, pattern = "^RP[SL]")
#What is the complexity of each cell library
gse141445.obj$log10GenesPerUMI <- log10(gse141445.obj$nFeature_RNA) / log10(gse141445.obj$nCount_RNA)

```

```{r}
# Summary statistics to guide threshold setting
cat("\nQC Metric Distributions:\n")
cat("nCount_RNA (UMI):\n")
cat("  Median:", median(gse141445.obj$nCount_RNA), 
    "| Q1-Q3:", quantile(gse141445.obj$nCount_RNA, 0.25), "-", 
    quantile(gse141445.obj$nCount_RNA, 0.75), "\n")
 
cat("nFeature_RNA (genes):\n")
cat("  Median:", median(gse141445.obj$nFeature_RNA),
    "| Q1-Q3:", quantile(gse141445.obj$nFeature_RNA, 0.25), "-",
    quantile(gse141445.obj$nFeature_RNA, 0.75), "\n")
 
cat("percent.mt:\n")
cat("  Median:", round(median(gse141445.obj$percent.mt), 2), "%",
    "| 95th percentile:", round(quantile(gse141445.obj$percent.mt, 0.95), 2), "%\n")
 
cat("percent.ribo:\n")
cat("  Median:", round(median(gse141445.obj$percent.ribo), 2), "%\n")
```

Violin Plot Distributions

```{r}
p4 <- VlnPlot(
  gse141445.obj,
  features =c("nCount_RNA", "nFeature_RNA", "percent.mt", "percent.ribo"),
  ncol = 4,
  pt.size = 0.1
) + theme(plot.title = element_text(face = "bold"))
p4

p5 <- FeatureScatter(gse141445.obj, feature1 = "nCount_RNA", feature2 = "nFeature_RNA") +
  labs(title = "UMI vs Genes Detected")
 
p6 <- FeatureScatter(gse141445.obj, feature1 = "nCount_RNA", feature2 = "percent.mt") +
  labs(title = "UMI vs Mitochondrial %")
 
p7 <- FeatureScatter(gse141445.obj, feature1 = "percent.mt", feature2 = "percent.ribo") +
  labs(title = "Mitochondrial % vs Ribosomal %")
 
p_scatter <- p5 + p6 + p7
p_scatter
```

No QC was necessary for GSE141445. Authors posted a clean version

```{r}
post_qc_rds <- "post_qc_rds_files"

if(!dir.exists(post_qc_rds)){
  dir.create(glue("./{post_qc_rds}"))
}

#no need for QC authors posted a post-QC version
saveRDS(gse141445.obj, "./post_qc_rds_files/gse141445.rds")
```

```{r}
gse141445.obj <- readRDS("./post_qc_rds_files/gse141445.rds")
```

```{r}
table(gse141445.obj@meta.data$PatientID)
```

# GSE176031/Song, H Download

```{r}
gse176031.obj <- readRDS("./sc_rds_files/gse176031.rds")
```

### Running Empty Drops

```{r}
object <- gse176031.obj
```

```{r}
DefaultAssay(gse176031.obj) <- "RNA"
empty_results <- emptyDrops_wrapr(gse176031.obj@assays$RNA@layers$counts, 
                                  object,
                                  lower = 100, 
                                  niters = 10000)
```

```{r}
object <- empty_results$subset_object
```

### Checking Sparsity

```{r}
#Calculate Matrix Sparsity
total_element_wise_counts <- (sum(object@assays$RNA@layers$counts) > 0)
dims <- (nrow(object) * ncol(object))
sparsity <- 1 - (total_element_wise_counts / dims)

cat("Matrix sparsity:", round(sparsity * 100, 1), "%\n")

#UMI distribution - bimodal obviously
umi_counts <- colSums(object@assays$RNA@layers$counts)
cat("Median UMI per droplet:", median(umi_counts), "\n")
cat("Droplets with >500 UMI:", sum(umi_counts > 500) , 
    "(likely cells, will be validated by EmptyDrops)\n")
```

### Checking Ambient RNA

```{r}
#Prepare SoupX channel
#tod (table of droplets) = raw counts from ALL droplets
#toc (table of cells) = filtered counts from validation cells only
sc <- SoupChannel(tod =  gse176031.obj@assays$RNA@layers$counts, 
                  toc = object@assays$RNA@layers$counts)
rownames(sc$metaData) <- colnames(object)
```

```{r}
DefaultAssay(object) <- "RNA"
#basic dimension reductions not set in stone because these are important
#For cluster annotation
object <- NormalizeData(object, verbose=F)
object <- FindVariableFeatures(object, nfeatures = 5000, verbose=F)
object <- ScaleData(object, verbose = FALSE)
object <- RunPCA(object, npcs=15, verbose = FALSE)
object <- FindNeighbors(object, dims = 1:15, verbose=F)
object <- FindClusters(object, resolution = 0.8, verbose =F)
object <- RunUMAP(object, dims = 1:15, verbose=F)



barcode_cluster <- setNames(as.character(object$seurat_clusters), colnames(object))
#setting clusters in the meta-data object
sc <- setClusters(sc, barcode_cluster)
```

```{r}

sc <- tryCatch({
  autoEstCont(sc, verbose = FALSE)
}, error = function(e) {
  cat("Note: autoEstCont failed\n")
  sc$fit$rho <- NULL
  return(sc)
})

contamination_fraction <- sc$fit$rho

# Check contamination level
cat("Contamination fraction:", 
    ifelse(is.null(contamination_fraction), "NULL (very clean sample)", 
           paste0(round(contamination_fraction * 100, 2), "%")), "\n")

# Decision: Skip or correct based on contamination level
if (is.null(contamination_fraction) || contamination_fraction < 0.05) {
  cat("→ Contamination is NULL or <5% - skipping SoupX correction\n")
  cat("   Your sample is clean! Proceeding with original counts.\n")
} else {
  cat("→ Contamination is", round(contamination_fraction * 100, 2), "% (≥5%)\n")
  cat("   Applying SoupX correction...\n")
  
  # Apply correction
  suppressWarnings({
    corrected_counts <- adjustCounts(sc)
  })
  object <- SetAssayData(object, layer = "counts", new.data = corrected_counts)
  cat("   ✓ SoupX correction applied\n")
}
```

### Checking Doublets

```{r}
#as.SingleCellExperiment expect one layer counts before adding
object <- JoinLayers(object)

sce <- as.SingleCellExperiment(object)

set.seed(42)
suppressWarnings({
  sce <- scDblFinder(sce, 
                     dbr = NULL, 
                     verbose = FALSE)
})

# Extract results
object$doublet_class <- ifelse(sce$scDblFinder.class == "doublet", "Doublet", "Singlet")
object$doublet_score <- sce$scDblFinder.score

#Tracking cells before doublet removal for QC summary
n_doublets <- sum(object$doublet_class == "Doublet")
doublet_rate <- n_doublets / ncol(object) * 100
cat("Doublets:", n_doublets, "(", round(doublet_rate, 1), "%)\n")

cells_before_doublet_removal <- ncol(object)
cat("Cells before doouble removal:", cells_before_doublet_removal)

p3 <- DimPlot(
  object,
  reduction = "umap",
  group.by = "doublet_class",
  cols = c("Singlet" = "#90E0EF", "Doublet" = "#EF233C"),
  pt.size = 1
) +
  labs(title = paste0("Doublet Detection (", round(doublet_rate, 1), "%)")) +
  theme(plot.title = element_text(face = "bold", size = 14))

p3
```

### Manual QC Review

Had in error occur in my single cell qc scatter plot function that blocked the calculation of percent mitochondrial gene percentage.The issue was that a group of cells was completely empty was causing a divide by 0 error. For some reason in the Rmd this it would be fine but on the sourced function this would cause an NA error to occur in the pt.mito column formed after using the Seurat function PercentageFeatureSet for \^MT-. The block below is to get rid of those cells completely from my Seurat object so that they don't cause further harm in future analysis

```{r}
# before applying empty droplet
QC_testing <- gse176031.obj
dataset_name <- "gse176031"
qc_phase <- "pre_thresholding"
#After applying empty droplets
#QC_testing <- object

#percentage of mitochondrial RNA counts
#calculate percentageof mitochondrial RNA per barcode
QC_testing[["percent.mt"]] <- PercentageFeatureSet(QC_testing, pattern = "^MT-")
#percentage of ribosomal RNA counts
QC_testing[["percent.ribo"]] <- PercentageFeatureSet(QC_testing, pattern = "^RP[SL]")
#What is the complexity of each cell library
QC_testing$log10GenesPerUMI <- log10(QC_testing$nFeature_RNA) / log10(QC_testing$nCount_RNA)
```

```{r}
length(colnames(gse176031.obj))
grep("^MT-", rownames(QC_testing), value = TRUE)

mt.genes <- grep("^MT-", rownames(QC_testing), value = TRUE)

mt.counts <- Matrix::colSums(
  GetAssayData(QC_testing, assay = DefaultAssay(QC_testing), layer = "counts")[mt.genes, , drop = FALSE]
)

summary(mt.counts)

na_cell_list <- names(mt.counts[is.na(QC_testing$percent.mt)])
na_cells <- rownames(gse176031.obj@meta.data[rownames(gse176031.obj@meta.data) %in% na_cell_list, ])
clean_cells <- colnames(gse176031.obj)[!colnames(gse176031.obj) %in% na_cells]

#keep the cells that are not
clean_gse176031.obj <- subset(gse176031.obj, cells = clean_cells)


# before applying empty droplet
QC_testing <- clean_gse176031.obj

length(colnames(QC_testing))
```

Pre-threshold Scatter Plots

```{r}
scqc_scatters(QC_testing, 
              dataset_name, 
              qc_phase)
```

Yes QC is most definitely necessary for this dataset. Although I am unsure of the cause currently. Is it because of native empty cells or was it method the Seurat Object was construcuted.

QC is required for this dataset

**What was found:**

25 cells are completelty empty, EmptyDropper returned 23,000 cells are considerd "empty" which makes no sense but we definitely know at least 30 are empty

### Cell Based Filtering

nfeature: represents the total number of unique genes detected in each individual cell

nCount: total sum of all UMI counts detected across all genes within that cell

```{r}
#
nfeature_min <- 200
  
nfeature_max <- 7500

ncount_min <- 500

ncount_max <- 55000
  
mt_thresh <- 10
```

```{r}
# before applying empty droplet
QC_testing <- clean_gse176031.obj

#percentage of mitochondrial RNA counts
#calculate percentageof mitochondrial RNA per barcode
QC_testing[["percent.mt"]] <- PercentageFeatureSet(QC_testing, pattern = "^MT-")
#percentage of ribosomal RNA counts
QC_testing[["percent.ribo"]] <- PercentageFeatureSet(QC_testing, pattern = "^RP[SL]")
#What is the complexity of each cell library
QC_testing$log10GenesPerUMI <- log10(QC_testing$nFeature_RNA) / log10(QC_testing$nCount_RNA)


scqc_prethreshold_check <-scqc_thresholds(QC_testing,
                                            "gse170631",
                                            nfeature_min,
                                            nfeature_max,
                                            ncount_min,
                                            mt_thresh)
```

```{r}
scqc_prethreshold_check
```

The reported post QC cell count was \~21,000 using all the samples (6 Prostate biopsies, 4 radical prostatectomies with tumor-only, 4 radical prostatectomies with matched normal samples from 4 patients (n=8), 7 Organoid samples derived from Patients). Safe to say their assays were not in good shape. Ideally post QC should have a lot more, but we can infer that the authors experiment did not go well.

**Current Thresholds:**

**nfeature_min** \<- 200: lenient threshold adjusted using **log10(UMI +1) vs log10(Genes + 1)**

**nfeature_max** \<- 7500: informed by initial scatter plot **(UMI vs Genes Detected)**

**ncount_min** \<- 600 lenient threshold adjusted using **log10(UMI +1) vs log10(Genes + 1)**

**ncount_max** \<- 75000: informed by initial scatter plot **(UMI vs Genes Detected)**

**mt_thresh** \<- 10, informed by **UMI vs Mitochondrial %** scatter plot

The reported post QC cell count was \~21,000 from the paper using all the samples (6 Prostate biopsies, 4 radical prostatectomies with tumor-only, 4 radical prostatectomies with matched normal samples from 4 patients (n=8), 7 Organoid samples derived from Patients). Safe to say their 10XChromium experiments were not in good shape. Ideally post QC should have a lot more.

### Subsetting After Filtering

```{r}
qc_df <- scqc_prethreshold_check$qc_df

pass_qc <- scqc_prethreshold_check$qc_df_pass
```

```{r}
cat("Set thresholds (adjust based on YOUR data):\n")
cat("  nFeature_RNA: [", nfeature_min, ",", nfeature_max, "]\n")
cat("  nCount_RNA: [", ncount_min, ",", ncount_max, "]\n")
cat("  percent.mt: <", mt_thresh, "%\n\n")

cat("Guidelines for threshold setting:\n")
cat("  • nFeature_RNA min: Where does the low-quality tail end? (typically 500-1000)\n")
cat("  • nFeature_RNA max: Where do potential doublets start? (typically 4000-7000)\n")
cat("  • percent.mt: PBMCs <10%, Brain <5%, Tumor <20%\n")
cat("  • If distributions are bimodal, set threshold between modes\n")
cat("  • Aim to remove no more than 25% of cells \n\n")

removal_pct <- sum(!qc_df$pass_qc) / nrow(qc_df) * 100
cat("Filtering impact:\n")
cat("  Cells before:", nrow(qc_df), "\n")
cat("  Cells passing QC:", sum(qc_df$pass_qc), "\n")
cat("  Cells removed:", sum(!qc_df$pass_qc), 
    "(", round(removal_pct, 1), "%)\n\n")

# Breakdown by criterion
cat("Removal breakdown:\n")
cat("  Low genes (<", nfeature_min, "):", 
    sum(qc_df$nFeature_RNA < nfeature_min), "\n")
cat("  High genes (>", nfeature_max, "):", 
    sum(qc_df$nFeature_RNA > nfeature_max), "\n")
cat("  High MT% (>", mt_thresh, "%):", 
    sum(qc_df$percent.mt > mt_thresh), "\n\n")

# Sanity check on removal rate
if (removal_pct > 30) {
  cat("⚠ WARNING: Removing", round(removal_pct, 1), "% of cells is high!\n")
  cat("   Consider relaxing thresholds (especially MT%)\n\n")
} else if (removal_pct < 5) {
  cat("⚠ WARNING: Only removing", round(removal_pct, 1), "% of cells is low.\n")
  cat("   Check if you're retaining low-quality cells.\n\n")
} else {
  cat("✓ Removal rate is reasonable (typical: 10-25%)\n\n")
}

# Track cells before cell-level QC filtering for summary
cells_before_cell_qc <- nrow(qc_df)

```

```{r}
gse176031.obj <- QC_testing
```

```{r}
post_qc_rds <- "post_qc_rds_files"

if(!dir.exists(post_qc_rds)){
  dir.create(glue("./{post_qc_rds}"))
}
```

```{r}
gse176031.obj <- subset(gse176031.obj,
                        subset = nCount_RNA >= ncount_min &
                          nCount_RNA <= ncount_max &
                          nFeature_RNA >= nfeature_min &
                          nFeature_RNA <= nfeature_max &
                          percent.mt < mt_thresh)


saveRDS(gse176031.obj, "./post_qc_rds_files/gse176031.rds")
```

**Check Post Threshold Scatter Plots**

```{r}
gse176031.obj
```

```{r}
dataset_name <- "176031"
qc_phase <- "post_threshold"

scqc_scatters(gse176031.obj, 
              dataset_name, 
              qc_phase)

scqc_postthreshold_check <- scqc_thresholds(gse176031.obj,
                                          "gse170631",
                                          "post_threshold",
                                          nfeature_min,
                                          nfeature_max,
                                          ncount_min,
                                          mt_thresh)
scqc_postthreshold_check
```

```{r}
gse176031.obj <- readRDS("./post_qc_rds_files/gse176031.rds")
gse176031.obj@meta.data
```

Checking on how many samples are within the datasets

```{r}
library(reticulate)
geo_source <- unique(gse176031.obj@meta.data$geo_source)
geo_source
r_to_py(geo_source)
```

```{python}
geo_source = r.geo_source
geo_source = [i.split("_", 1)[1] for i in geo_source]
sample_dict = {}
for i in geo_source:
  if i not in sample_dict:
    sample_dict[i] = 1
  else:
    sample_dict[i] += 1

sorted_dict = {key: value for key, value in sorted(sample_dict.items())}
print(sorted_dict)
nums = [val for val in sorted_dict.values()]
print(*nums, sep = " ")
```

# [GSE150692 (Crowley, 2020) (PRAD](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE150692)) Download

```{r}
gse150692.obj <- readRDS("./sc_rds_files/gse150692.rds")
```

### Running Empty Drops

```{r}
object <- gse150692.obj
```

```{r}
empty_results <- emptyDrops_wrapr(gse150692.obj@assays$RNA@layers$counts, 
                                  object,
                                  lower = 100, 
                                  niters = 10000)
```

### Checking Sparsity

```{r}
sum(object@assays$RNA@layers$counts)
sum(is.na(object@assays$RNA@layers$counts))
```

```{r}
#Calculate Matrix Sparsity
total_element_wise_counts <- (sum(object@assays$RNA@layers$counts) > 0)
print(total_element_wise_counts)
dims <- (nrow(object) * ncol(object))
sparsity <- 1 - (total_element_wise_counts / dims)

cat("Matrix sparsity:", round(sparsity * 100, 1), "%\n")

#UMI distribution - bimodal obviously
umi_counts <- colSums(object@assays$RNA@layers$counts)
cat("Median UMI per droplet:", median(umi_counts), "\n")
cat("Droplets with >500 UMI:", sum(umi_counts > 500) , 
    "(likely cells, will be validated by EmptyDrops)\n")
```

```{r}
object@assays$RNA@cells
```

```{r}
length(colnames(object))
length(rownames(object@meta.data))
identical(colnames(object), rownames(object@meta.data))
setequal(colnames(object), rownames(object@meta.data))


length(colnames(object[["RNA"]]))
nrow(object@meta.data)

identical(colnames(object[["RNA"]]), rownames(object@meta.data))
```

### Checking Ambient RNA

```{r}
#Prepare SoupX channel
#tod (table of droplets) = raw counts from ALL droplets
#toc (table of cells) = filtered counts from validation cells only
sc <- SoupChannel(tod = object@assays$RNA$counts, 
                  toc = empty_results)
rownames(sc$metaData) <- colnames(object)
```

```{r}
DefaultAssay(object) <- "RNA"
#basic dimension reductions not set in stone because these are important
#For cluster annotation
object <- NormalizeData(object, verbose=F)
object <- FindVariableFeatures(object, nfeatures = 5000, verbose=F)
object <- ScaleData(object, verbose = FALSE)
object <- RunPCA(object, npcs=15, verbose = FALSE)
object <- FindNeighbors(object, dims = 1:15, verbose=F)
object <- FindClusters(object, resolution = 0.8, verbose =F)
object <- RunUMAP(object, dims = 1:15, verbose=F)



barcode_cluster <- setNames(as.character(object$seurat_clusters), colnames(object))
#setting clusters in the meta-data object
sc <- setClusters(sc, barcode_cluster)
```

```{r}

sc <- tryCatch({
  autoEstCont(sc, verbose = FALSE)
}, error = function(e) {
  cat("Note: autoEstCont failed\n")
  sc$fit$rho <- NULL
  return(sc)
})

contamination_fraction <- sc$fit$rho

# Check contamination level
cat("Contamination fraction:", 
    ifelse(is.null(contamination_fraction), "NULL (very clean sample)", 
           paste0(round(contamination_fraction * 100, 2), "%")), "\n")

# Decision: Skip or correct based on contamination level
if (is.null(contamination_fraction) || contamination_fraction < 0.05) {
  cat("→ Contamination is NULL or <5% - skipping SoupX correction\n")
  cat("   Your sample is clean! Proceeding with original counts.\n")
} else {
  cat("→ Contamination is", round(contamination_fraction * 100, 2), "% (≥5%)\n")
  cat("   Applying SoupX correction...\n")
  
  # Apply correction
  suppressWarnings({
    corrected_counts <- adjustCounts(sc)
  })
  object <- SetAssayData(object, layer = "counts", new.data = corrected_counts)
  cat("   ✓ SoupX correction applied\n")
}
```

### Checking Doublets

```{r}
object <- JoinLayers(object)

sce <- as.SingleCellExperiment(object)

set.seed(42)
suppressWarnings({
  sce <- scDblFinder(sce, 
                     dbr = NULL, 
                     verbose = FALSE)
})

# Extract results
object$doublet_class <- ifelse(sce$scDblFinder.class == "doublet", "Doublet", "Singlet")
object$doublet_score <- sce$scDblFinder.score

#Tracking cells before doublet removal for QC summary
n_doublets <- sum(object$doublet_class == "Doublet")
doublet_rate <- n_doublets / ncol(object) * 100
cat("Doublets:", n_doublets, "(", round(doublet_rate, 1), "%)\n")

cells_before_doublet_removal <- ncol(object)
cat("Cells before doouble removal:", cells_before_doublet_removal)

p3 <- DimPlot(
  object,
  reduction = "umap",
  group.by = "doublet_class",
  cols = c("Singlet" = "#90E0EF", "Doublet" = "#EF233C"),
  pt.size = 1
) +
  labs(title = paste0("Doublet Detection (", round(doublet_rate, 1), "%)")) +
  theme(plot.title = element_text(face = "bold", size = 14))

p3
```

### Manual Review

```{r}
# before applying empty droplet
QC_testing <- gse150692.obj

#After applying empty droplets
#QC_testing <- object


# Calculate QC metrics

#percentage of mitochondrial RNA counts
#calculate percentageof mitochondrial RNA per barcode
QC_testing[["percent.mt"]] <- PercentageFeatureSet(QC_testing, pattern = "^MT-")
#percentage of ribosomal RNA counts
QC_testing[["percent.ribo"]] <- PercentageFeatureSet(QC_testing, pattern = "^RP[SL]")
#What is the complexity of each cell library
QC_testing$log10GenesPerUMI <- log10(QC_testing$nFeature_RNA) / log10(QC_testing$nCount_RNA)
```

```{r}
dataset_name <- "gse150692"
qc_phase <- "pre_threshold"
scqc_scatters <- scqc_scatters(QC_testing, 
              dataset_name, 
              qc_phase)

```

```{r}

```

**What was found ?**

As advertised within the data name these are already QCed by the authors. However it's concernering to see droplets with both low nFeature and nCounts. This could mean either these are dying cells that are only showing mitochondrial transcripts or that they contain cells with total low unique transcript expressions.

### Cell Based Filtering Thresholds

**nfeature:** represents the total number of unique genes detected in each individual cell

**nCount:** total sum of all UMI counts detected across all genes within that cell

```{r}
# minimal thresholding to these

nfeature_min <- 200

nfeature_max <- 8000

ncount_min <- 1000

ncount_max <- 50000

mt_thresh <- 10

scqc_prethreshold_check <-scqc_thresholds(QC_testing,
                                          dataset_name,
                                          qc_phase = qc_phase,
                                          nfeature_min,
                                          nfeature_max,
                                          ncount_min,
                                          mt_thresh)
scqc_prethreshold_check
```

Due to low Doublet Report (6.9%), and non-return of EmptyDrolets Wrapper function. I've decided that applying QC thresholds to this dataset would be too strict relative to metrics (below), and graphs that suggest that it already has been cleaned by the authors and is even stated within their GEOquery submission that it has been cleaned. Applying manual thresholds runs the risk of cutting relevant signal.

```         
nCount_RNA (UMI):
  Median: 6225 | Q1-Q3: 4353 - 9421.25 
nFeature_RNA (genes):
  Median: 2488 | Q1-Q3: 1929.5 - 3194.25 
percent.mt:
  Median: 5.9 % | 95th percentile: 13.55 %
percent.ribo:
  Median: 0 %
```

```{r}

post_qc_rds <- "post_qc_rds_files"

if(!dir.exists(post_qc_rds)){
  dir.create(glue("./{post_qc_rds}"))
}

saveRDS(gse150692.obj, "./post_qc_rds_files/gse150692.rds")
```

```{r}
gse150692.obj <- readRDS("./post_qc_rds_files/gse150692.rds")
```

```{r}
gse150692.obj@meta.data$
```

# GSE181294 ([Chen et. al., 2023](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE181294))

```{r}
gse181294.obj <- readRDS("./sc_rds_files/gse181294.rds")
```

```{r}
gse181294.obj
object <- gse181294.obj
#based on the initial cell counts we will have
# empty droplets
```

```{r}
summary(rowSums(gse181294.obj@assays$RNA@layers$counts))
```

### Running Empty Droplets

```{r}
empty_results <- emptyDrops_wrapr(gse181294.obj@assays$RNA@layers$counts, 
                                  object,
                                  lower = 100, 
                                  niters = 10000)
```

### Checking Sparsity

```{r}
#Calculate Matrix Sparsity
total_element_wise_counts <- (sum(object@assays$RNA@layers$counts) > 0)
dims <- (nrow(object) * ncol(object))
sparsity <- 1 - (total_element_wise_counts / dims)

cat("Matrix sparsity:", round(sparsity * 100, 1), "%\n")

#UMI distribution - bimodal obviously
umi_counts <- colSums(object@assays$RNA@layers$counts)
cat("Median UMI per droplet:", median(umi_counts), "\n")
cat("Droplets with >500 UMI:", sum(umi_counts > 500) , 
    "(likely cells, will be validated by EmptyDrops)\n")
```

Verifiying that all barcodes are

```{r}
length(colnames(object))
length(rownames(object@meta.data))
identical(colnames(object), rownames(object@meta.data))
setequal(colnames(object), rownames(object@meta.data))


length(colnames(object[["RNA"]]))
nrow(object@meta.data)

identical(colnames(object[["RNA"]]), rownames(object@meta.data))
```

### Setup Channel for SoupX

Preparing Soup Channel

```{r}
#Prepare SoupX channel
#tod (table of droplets) = raw counts from ALL droplets
#toc (table of cells) = filtered counts from validation cells only
sc <- SoupChannel(tod = object@assays$RNA$counts, 
                  toc = empty_results)
rownames(sc$metaData) <- colnames(object)
```

###Ambient RNA checking

Setting up Temporary Object

```{r}
#basic dimension reductions not set in stone because these are important
#For cluster annotation
object <- NormalizeData(object, verbose=F)
object <- FindVariableFeatures(object, nfeatures = 5000, verbose=F)
object <- ScaleData(object, verbose = FALSE)
object <- RunPCA(object, npcs=15, verbose = FALSE)
object <- FindNeighbors(object, dims = 1:15, verbose=F)
object <- FindClusters(object, resolution = 0.8, verbose =F)
object <- RunUMAP(object, dims = 1:15, verbose=F)
```

```{r}
barcode_cluster <- setNames(as.character(object$seurat_clusters), colnames(object))
#setting clusters in the meta-data object
sc <- setClusters(sc, barcode_cluster)
```

###Estimating Contanimation

```{r}
sc <- tryCatch({
  autoEstCont(sc, verbose = FALSE)
}, error = function(e) {
  cat("Note: autoEstCont failed\n")
  sc$fit$rho <- NULL
  return(sc)
})

contamination_fraction <- sc$fit$rho

# Check contamination level
cat("Contamination fraction:", 
    ifelse(is.null(contamination_fraction), "NULL (very clean sample)", 
           paste0(round(contamination_fraction * 100, 2), "%")), "\n")
 
# Decision: Skip or correct based on contamination level
if (is.null(contamination_fraction) || contamination_fraction < 0.05) {
  cat("→ Contamination is NULL or <5% - skipping SoupX correction\n")
  cat("   Your sample is clean! Proceeding with original counts.\n")
} else {
  cat("→ Contamination is", round(contamination_fraction * 100, 2), "% (≥5%)\n")
  cat("   Applying SoupX correction...\n")
 
  # Apply correction
  suppressWarnings({
    corrected_counts <- adjustCounts(sc)
  })
  object <- SetAssayData(object, layer = "counts", new.data = corrected_counts)
  cat("   ✓ SoupX correction applied\n")
}
```

### Checking Doublets

```{r}
object <- JoinLayers(object)

sce <- as.SingleCellExperiment(object)

set.seed(42)
suppressWarnings({
  sce <- scDblFinder(sce, 
                     dbr = NULL, 
                     verbose = FALSE)
})

# Extract results
object$doublet_class <- ifelse(sce$scDblFinder.class == "doublet", "Doublet", "Singlet")
object$doublet_score <- sce$scDblFinder.score

#Tracking cells before doublet removal for QC summary
n_doublets <- sum(object$doublet_class == "Doublet")
doublet_rate <- n_doublets / ncol(object) * 100
cat("Doublets:", n_doublets, "(", round(doublet_rate, 1), "%)\n")

cells_before_doublet_removal <- ncol(object)
cat("Cells before doouble removal:", cells_before_doublet_removal)

p3 <- DimPlot(
  object,
  reduction = "umap",
  group.by = "doublet_class",
  cols = c("Singlet" = "#90E0EF", "Doublet" = "#EF233C"),
  pt.size = 1
) +
  labs(title = paste0("Doublet Detection (", round(doublet_rate, 1), "%)")) +
  theme(plot.title = element_text(face = "bold", size = 14))

p3
```

Doublet Detection was 31.6% Empty Droplets did not work. Investigate that tomorrow and summarize the QC process

### Manual Review

```{r}
# before applying empty droplet
#QC_testing <- gse150692.obj

#After applying empty droplets
QC_testing <- object


# Calculate QC metrics

#percentage of mitochondrial RNA counts
#calculate percentageof mitochondrial RNA per barcode
QC_testing[["percent.mt"]] <- PercentageFeatureSet(QC_testing, pattern = "^MT-")
#percentage of ribosomal RNA counts
QC_testing[["percent.ribo"]] <- PercentageFeatureSet(QC_testing, pattern = "^RP[SL]")
#What is the complexity of each cell library
QC_testing$log10GenesPerUMI <- log10(QC_testing$nFeature_RNA) / log10(QC_testing$nCount_RNA)
```

```{r}
dataset_name <- "gse181294"
qc_phase <- "pre_threshold"
scqc_scatters <- scqc_scatters(QC_testing, 
                               dataset_name, 
                               qc_phase)
```

### Cell Based Filtering Thresholds

**nfeature:** represents the total number of unique genes detected in each individual cell

**nCount:** total sum of all UMI counts detected across all genes within that cell

```{r}
# minimal thresholding to these

nfeature_min <- 500

nfeature_max <- 8000

ncount_min <- 1000

ncount_max <- 70000

mt_thresh <- 5

qc_phase <- "pre_threshold"

scqc_prethreshold_check <-scqc_thresholds(gse181294.obj,
                                          dataset_name,
                                          qc_phase = qc_phase,
                                          nfeature_min,
                                          nfeature_max,
                                          ncount_min,
                                          mt_thresh)
scqc_prethreshold_check
```

Keeping 90% of the Cells

nfeature_min \<- 500

nfeature_max \<- 8000

ncount_min \<- 1000

ncount_max \<- 70000

mt_thresh \<- 5

### Applying Thresholds

```{r}
gse181294.obj <- subset(gse181294.obj,
                        subset = nCount_RNA >= ncount_min &
                          nCount_RNA <= ncount_max &
                          nFeature_RNA >= nfeature_min &
                          nFeature_RNA <= nfeature_max &
                          percent.mt < mt_thresh)


saveRDS(gse181294.obj, "./post_qc_rds_files/gse181294.rds")
```

```{r}
nfeature_min <- 500

nfeature_max <- 8000

ncount_min <- 1000

ncount_max <- 70000

mt_thresh <- 5

dataset_name <- "gse181294"
qc_phase <- "post_threshold"
scqc_scatters <- scqc_scatters(gse181294.obj, 
                               dataset_name, 
                               qc_phase)

scqc_postthreshold_check <-scqc_thresholds(gse181294.obj,
                                          dataset_name,
                                          qc_phase = qc_phase,
                                          nfeature_min,
                                          nfeature_max,
                                          ncount_min,
                                          mt_thresh)
```

```{r}
scqc_scatters
```

```{r}
gse181294.obj <- readRDS("./post_qc_rds_files/gse181294.rds")
gse181294.obj
```
