---
title: "SC_QualityControl"
output: html_document
---

# Library Read In

```{r}
#Seurat Core
library(Seurat)
#library(BPCells)
library(SeuratWrappers)
library(S4Vectors)

#non-Saitja QC libraries
library(DropletUtils)
library(scDblFinder)
library(SoupX)
library(SingleCellExperiment)

#Visualizations
library(ggplot2)
library(patchwork)
library(dplyr)
library(tidyverse)
library(glue)
```

```{r}
overall_dir <- "qc_images"

if(!dir.exists(overall_dir)){
  dir.create(glue("./{overall_dir}"))
}
```

# GSE141445/Chen, O QC

```{r}
gse141445.obj <- readRDS("./sc_rds_files/gse141445.rds")
```

Estimate Sparsity

```{r}
object <- gse141445.obj
```

```{r}
#Calculate Matrix Sparsity
total_element_wise_counts <- (sum(object@assays$RNA@layers$counts) > 0)
dims <- (nrow(object) * ncol(object))
sparsity <- 1 - (total_element_wise_counts / dims)

cat("Matrix sparsity:", round(sparsity * 100, 1), "%\n")

#UMI distribution - bimodal obviously
umi_counts <- colSums(object@assays$RNA@layers$counts)
cat("Median UMI per droplet:", median(umi_counts), "\n")
cat("Droplets with >500 UMI:", sum(umi_counts > 500) , 
    "(likely cells, will be validated by EmptyDrops)\n")
```

### Empty Droplet Detection

Need to write write description of the empty Droplets functions.

```{r}
emptyDrops_wrapr <- function(rna_mat, object) {
  #use a local variable to save the original matrix in here
  original_mat <- rna_mat
  
  initial_cell_cnt <- dim(original_mat)[1]
  initial_gene_cnt <- dim(original_mat)[2]
  
  #start of tryCatch
  tryCatch(
    {
      set.seed(100)
      #lower: Droplets with <100 UMI assumed empty. Depending on the experiment this could change
      #niters: Iterations for specifying the number of iterations to use for the Monte Carlo p-value calculations.
      empty_report <- emptyDrops(
        m = rna_mat,
        lower = 100,
        niters = 10000,
        test.ambient = TRUE
      )
      
      #filtering results
      #Identify cell with FDR < 0.01
      is_cell <- empty_report$FDR < 0.01
      is_cell[is.na(is_cell)] <- FALSE # Treat NAs as empty (very low UMIs per drolet)
      validated_barcodes <- colnames(object)[is_cell] # filtering by validated cells
      
      
      cat("Droplets tested:", ncol(object), "\n")
      cat("Cells called:", sum(is_cell), 
          "(", round(sum(is_cell)/ncol(object)*100, 1), "%)\n")
      cat("Empty droplets removed:", sum(!is_cell), 
          "(", round(sum(!is_cell)/ncol(object)*100, 1), "%)\n")
      
      
      #
      save_raw_counts <- original_mat
      
      #filtering the Seurat Object by the validated BardCodes
      new_object <- subset(object, cells = validated_barcodes)
      
      cat("After Empty Drops:", ncol(new_object), "cells retained\n")
      
      
      return(list(empty_droplets, new_object))
    },
    #tryCatch error function will return the original matrix
    #This part is necessary because the author of the
    error = function(e) {
      message("Matrix Already Filtered by CellRanger")
      return(original_mat)
    }#end of error bracket
      
  )# end of tryCatch
    
  
}# end of wrapr function
```

```{r}
empty_results <- emptyDrops_wrapr(object@assays$RNA@layers$counts, object)
```

```{r}
empty_results
```

### Setup Channel for SoupX

Preparing Soup Channel

```{r}
#Prepare SoupX channel
#tod (table of droplets) = raw counts from ALL droplets
#toc (table of cells) = filtered counts from validation cells only
sc <- SoupChannel(tod = object@assays$RNA$counts, 
                  toc = empty_results)
rownames(sc$metaData) <- colnames(object)
```

Ambient RNA checking

```{r}
#basic dimension reductions not set in stone because these are important
#For cluster annotation
object <- NormalizeData(object, verbose=F)
object <- FindVariableFeatures(object, nfeatures = 5000, verbose=F)
object <- ScaleData(object, verbose = FALSE)
object <- RunPCA(object, npcs=15, verbose = FALSE)
object <- FindNeighbors(object, dims = 1:15, verbose=F)
object <- FindClusters(object, resolution = 0.8, verbose =F)
object <- RunUMAP(object, dims = 1:15, verbose=F)
```

```{r}
barcode_cluster <- setNames(as.character(object$seurat_clusters), colnames(object))
#setting clusters in the meta-data object
sc <- setClusters(sc, barcode_cluster)
```

Estimating Contanimation

```{r}
sc <- tryCatch({
  autoEstCont(sc, verbose = FALSE)
}, error = function(e) {
  cat("Note: autoEstCont failed\n")
  sc$fit$rho <- NULL
  return(sc)
})

contamination_fraction <- sc$fit$rho

# Check contamination level
cat("Contamination fraction:", 
    ifelse(is.null(contamination_fraction), "NULL (very clean sample)", 
           paste0(round(contamination_fraction * 100, 2), "%")), "\n")
 
# Decision: Skip or correct based on contamination level
if (is.null(contamination_fraction) || contamination_fraction < 0.05) {
  cat("→ Contamination is NULL or <5% - skipping SoupX correction\n")
  cat("   Your sample is clean! Proceeding with original counts.\n")
} else {
  cat("→ Contamination is", round(contamination_fraction * 100, 2), "% (≥5%)\n")
  cat("   Applying SoupX correction...\n")
 
  # Apply correction
  suppressWarnings({
    corrected_counts <- adjustCounts(sc)
  })
  object <- SetAssayData(object, layer = "counts", new.data = corrected_counts)
  cat("   ✓ SoupX correction applied\n")
}
```

Doublet Detection Doublets are when two cell or nucei are contained within a droplet after partition instead of the intended co-encapsulation 10x CellRanger captures \~1% doublets per 1000 cell load When you have a higher loading you'll likely have more doublets (10,000 cells \~6% - 9% doublets)

Rates \>10% suggest problems with cell loading concentration or the doublet detector being too sensitive. Given that our example sample includes over 9,800 cells, a 10% rate is considered acceptable.

**This can create false "hybrid" cell types which can**

1\. Inflate Cell counts

2\. Confound trajectory analysis

3\. Mess up differential expression

```{r}
object <- JoinLayers(object)

sce <- as.SingleCellExperiment(object)

set.seed(42)
suppressWarnings({
  sce <- scDblFinder(sce, 
                     dbr = NULL, 
                     verbose = FALSE,
                     clusters = TRUE)
})

# Extract results
object$doublet_class <- ifelse(sce$scDblFinder.class == "doublet", "Doublet", "Singlet")
object$doublet_score <- sce$scDblFinder.score

#Tracking cells before doublet removal for QC summary
n_doublets <- sum(object$doublet_class == "Doublet")
doublet_rate <- n_doublets / ncol(object) * 100
cat("Doublets:", n_doublets, "(", round(doublet_rate, 1), "%)\n")

cells_before_doublet_removal <- ncol(object)
cat("Cells before doouble removal:", cells_before_doublet_removal)

p3 <- DimPlot(
  object,
  reduction = "umap",
  group.by = "doublet_class",
  cols = c("Singlet" = "#90E0EF", "Doublet" = "#EF233C"),
  pt.size = 1
) +
  labs(title = paste0("Doublet Detection (", round(doublet_rate, 1), "%)")) +
  theme(plot.title = element_text(face = "bold", size = 14))
```

```{r}
table(truth=sce$type, call=sce$scDblFinder.class)
```

```{r}
ggsave("qc_images/02_doublets_umap.png", p3, width = 8, height = 7, dpi = 300)

object <- subset(object, subset = doublet_class == "Singlet")

cat("After doublet removal:", ncol(object), "cells\n")
```

### manual thresholding for Cell level quality

```{r}
# Calculate QC metrics

#percentage of mitochondrial RNA counts
#calculate percentageof mitochondrial RNA per barcode
gse141445.obj[["percent.mt"]] <- PercentageFeatureSet(gse141445.obj, pattern = "^MT-")
#percentage of ribosomal RNA counts
gse141445.obj[["percent.ribo"]] <- PercentageFeatureSet(gse141445.obj, pattern = "^RP[SL]")
#What is the complexity of each cell library
gse141445.obj$log10GenesPerUMI <- log10(gse141445.obj$nFeature_RNA) / log10(gse141445.obj$nCount_RNA)

```

```{r}
# Summary statistics to guide threshold setting
cat("\nQC Metric Distributions:\n")
cat("nCount_RNA (UMI):\n")
cat("  Median:", median(gse141445.obj$nCount_RNA), 
    "| Q1-Q3:", quantile(gse141445.obj$nCount_RNA, 0.25), "-", 
    quantile(gse141445.obj$nCount_RNA, 0.75), "\n")
 
cat("nFeature_RNA (genes):\n")
cat("  Median:", median(gse141445.obj$nFeature_RNA),
    "| Q1-Q3:", quantile(gse141445.obj$nFeature_RNA, 0.25), "-",
    quantile(gse141445.obj$nFeature_RNA, 0.75), "\n")
 
cat("percent.mt:\n")
cat("  Median:", round(median(gse141445.obj$percent.mt), 2), "%",
    "| 95th percentile:", round(quantile(gse141445.obj$percent.mt, 0.95), 2), "%\n")
 
cat("percent.ribo:\n")
cat("  Median:", round(median(gse141445.obj$percent.ribo), 2), "%\n")
```

Violin Plot Distributions

```{r}
p4 <- VlnPlot(
  gse141445.obj,
  features =c("nCount_RNA", "nFeature_RNA", "percent.mt", "percent.ribo"),
  ncol = 4,
  pt.size = 0.1
) + theme(plot.title = element_text(face = "bold"))
p4

p5 <- FeatureScatter(gse141445.obj, feature1 = "nCount_RNA", feature2 = "nFeature_RNA") +
  labs(title = "UMI vs Genes Detected")
 
p6 <- FeatureScatter(gse141445.obj, feature1 = "nCount_RNA", feature2 = "percent.mt") +
  labs(title = "UMI vs Mitochondrial %")
 
p7 <- FeatureScatter(gse141445.obj, feature1 = "percent.mt", feature2 = "percent.ribo") +
  labs(title = "Mitochondrial % vs Ribosomal %")
 
p_scatter <- p5 + p6 + p7
p_scatter
```

*QC Strategy*

Manual QC First when plotting Violin plots and scatterplots look for Bimodal distributions Outliers in scatter plot Reltionships between metrics

Hierarchy of thresholds

nFeature_RNA (set threshold) to catch doublets or multiplets above the high threshold and empty droplets on the lower threshold

*nfeature_min / nfeature_max*: set thresholds for the set of cells that contain above a certain number of feature per recorded library (barcode).The purpose of he lower part will exclude potential empty or ambient RNA droplets While the maximum threshold's purpose is to exclude potenital doublets with a lot more unnecessart and possible assessment changing information of UMIs within 1 droplet but is recorded on 1 barcode.

Common thresholds(Check the visuals always)

nfeature_max = 5000

nfeature_min = 500

*ncount_min / ncount_max*: thresholds for the total UMIs within 1 column to determine what a co-encapsulated cell look within the data. This will set threshold at the nCount_RNA data

Common thresholds

ncount_min = 800

ncount_max = 20,000

nCount_RNA vs nFeature_RNA: Helps us to determine the complexity of a cell library *UMI vs Genes Detected (nCount_RNA vs nFeature_RNA)* Poor quality cells are likely to have low genes and UMIs per cell. Good cells/droplets exhbit both higher number of genes per cell and higher numbers of UMIs.

Good Combinations high UMI + high Feature Cnt UMI = high Feature Cnt Bad Combinatons extremely high UMI + extremely high Feature Cnt

*mt-thresh*: Determines what health level the "cells" should be within the droplet. Sometime cells could be in the process of dying out and thus the barcodes will capture mRNA from mitochondria. Ideal is \<5% - 10%. However this is dependent on the purpose of the study. If you're studying DYING CELLs keep it high.

ribosomal-RNA%: High ribosomal percentage is not always "bad". I could indicate activity or specific cell types like plasma stem cells, but trasking it help to detect technical batch effect or hyper-transscribing cell states

Novelty Score ( log10(nFeature)/ log10(nCout))

Captures the ratio of unique genes detected relative to total sequencing depth. Good quality cells generally have values \> 0.80 . Values below $0.70$–$0.75$ indicate low-complexity libraries (e.g., a few overrepresented transcripts, highly degraded samples, or high contamination).

Resort Automatic Strategies AFTER you've completed the manual Find EmptyDroplets (DropletUtils) Find AmbientRNA (SoupX) Find Doublets (DoubletFinder)
